\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{bm}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multirow}
\usepackage{array}
\usepackage{enumitem}
\usepackage{parskip}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric,arrows.meta,positioning,fit,backgrounds,calc}

\geometry{margin=1in, top=0.9in, bottom=0.9in}

% NeurIPS-style: tighter spacing
\setlength{\parskip}{3pt}

\definecolor{codeblue}{RGB}{30,100,200}
\definecolor{layerA}{RGB}{210,230,255}
\definecolor{layerB}{RGB}{210,255,220}
\definecolor{layerC}{RGB}{245,220,255}
\definecolor{layerD}{RGB}{255,235,210}
\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{codeblue}\bfseries,
  commentstyle=\color{gray},
  breaklines=true,
  frame=single
}

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

% -----------------------------------------------------------------------
\title{\textbf{Two Brains, One Network: Dual-Process Topology Management\\for Heterogeneous Physics-Informed Kolmogorov-Arnold Networks}}
\author{
  Richard Liu \quad Yufei Zou \\
  School of Data Science \\
  University of North Carolina at Charlotte \\
  \texttt{github.com/RicardoLaMo/OpKAN}
}
\date{March 2026}

\begin{document}
\maketitle

% -----------------------------------------------------------------------
% CHANGE 1: New abstract
\begin{abstract}
Physics-informed Kolmogorov-Arnold Networks (KANs) introduce a structural management
problem that gradient descent cannot solve: when should an edge remain a B-spline, be
replaced by a symbolic function, or be pruned? We formalize this as the \emph{topology
surgery problem} and show that it creates five chronic failure modes in heterogeneous
KANs: topology stagnation, parameterization lock-in, symbolic-numerical tension, regime
blindness, and instability in the second-derivative smoothness required by the PDE. To
address this, we propose a \textbf{dual-process topology agent}: a fast reflexive
controller performs high-frequency local maintenance such as pruning and learning-rate
stabilization, while a slow strategic controller performs low-frequency topology surgery
guided by PDE geometry and market-regime signals. A typed DSL, static AST allowlist, and
atomic rollback ensure that every mutation preserves twice-differentiable smoothness
($C^2$ continuity) and transactional integrity. We demonstrate the framework on a Heston
stochastic-volatility PI-KAN, where stable second derivatives are a hard requirement
because of the cross-derivative term. An ablation shows that System~2 is required for
regime-ordered behavior, while System~1 restores full throughput; PDE residual
minimization alone fails all five financial monotonicity checks under distributional
shift, revealing a broader training-objective insufficiency in physics-constrained
learning under regime change. The system sustains $\mathbf{26{,}300}$\,smp/sec on H200
with zero GPU blocking, showing that online structural governance need not compromise
training efficiency.
\end{abstract}

% =====================================================================
\section{Introduction}
\label{sec:intro}
% =====================================================================

\subsection{The Topology Surgery Problem in Heterogeneous KANs}
\label{sec:intro:surgery}

A heterogeneous KAN, where some edges remain B-splines, others are replaced by
closed-form symbolic functions, and others are pruned to zero, is not just a larger
MLP; it is a mutable mathematical object whose structure can change during training.

In such a network, edge parameterization is not a one-time hyperparameter choice
fixed before training. It is an ongoing structural decision that depends on evolving
activation profiles, regime signals, and available domain knowledge. Gradient descent
can optimize the coefficients of a B-spline edge, but it cannot decide that the edge
should switch class: from spline to symbolic form, or from a weak edge to zero. It
can drive a B-spline's coefficients toward zero, but provides no pruning signal
because the gradient of a near-zero activation is also near-zero. This is the
\emph{topology surgery problem}: a class of structural decisions that falls strictly
outside the scope of gradient-based optimization.

This problem manifests in five distinct chronic failure modes of heterogeneous KAN
topology management. Each failure mode arises from a distinct gradient-signal gap:
a setting in which the training loss provides no corrective signal for the structural
decision that must be made (Table~\ref{tab:problems}).

\begin{table}[ht]
\centering
\caption{Five chronic failure modes of heterogeneous KAN topology. Each arises from
a gradient-signal gap: the training loss provides no corrective signal for the
structural decision in question.}
\begin{tabular}{@{}p{3.2cm}p{4.5cm}p{4.5cm}@{}}
\toprule
Failure Mode & Gradient-Signal Gap & Observable Symptom \\
\midrule
Topology stagnation & $\nabla_\theta \mathcal{L} \approx 0$ when $\|\phi_e\|_1 \approx 0$ & Near-zero edges persist indefinitely \\
Parameterization lock-in & No gradient over $\sigma: E \to \mathcal{C}$ & B-splines approximating known functions remain B-splines \\
Symbolic-numerical tension & Symbolic edges have no $\theta$; gradient dead zones form & Neighboring B-splines underfit \\
Regime blindness & $\mathcal{L}_\text{PDE}$ is regime-agnostic & Regime-ordering violations under distributional shift \\
$C^2$ instability & No explicit $C^2$-norm term in $\mathcal{L}$ & Ill-conditioned cross-derivatives under high lr \\
\bottomrule
\end{tabular}
\label{tab:problems}
\end{table}

\begin{enumerate}[nosep]
  \item \textbf{Topology stagnation.} Near-zero B-spline edges persist through training
  because their loss gradient is also near-zero. The edge contributes nothing but
  continues consuming computation.

  \item \textbf{Parameterization lock-in.} Gradient descent never has incentive to
  change an edge's parameterization class. A B-spline that closely approximates
  $\sin(x)$ remains a B-spline indefinitely, losing both interpretability and
  efficiency.

  \item \textbf{Symbolic-numerical tension.} Symbolic edges have no trainable
  parameters. In a mixed topology, gradient cannot flow through symbolic edges to
  inform neighboring B-spline adaptations, creating dead zones in the optimization
  landscape.

  \item \textbf{Regime blindness.} Under distributional shift (market regime change),
  a frozen topology cannot adapt. Weight updates shift the learned function but cannot
  restructure which edges carry which mathematical primitives.

  \item \textbf{$C^2$ instability.} Aggressive learning-rate updates can make an edge
  insufficiently smooth for stable second-derivative computation. In the Heston
  setting this matters directly: the PDE contains a cross-derivative term
  $\rho\sigma vS\,\partial^2 V/\partial S\partial v$ that is evaluated by autograd,
  so unstable edge curvature propagates as NaN through the cross-derivative
  computation. The symptom is detectable by monitoring the L2 norm of second-derivative
  outputs during training.
\end{enumerate}

\subsection{Why Two Brains}
\label{sec:intro:twobrains}

The five problems decompose naturally by frequency and cognitive type. Problems~1
and~5, topology stagnation and second-derivative instability, are high-frequency and
low-complexity. They can be detected from local statistics such as edge activation
magnitude or second-derivative size, and handled with simple actions such as pruning
or learning-rate adjustment. These are reflexive maintenance tasks; a fast model
($<$50\,ms, 7B parameters) is sufficient.

Problems~2, 3, and~4, parameterization lock-in, symbolic-numerical tension, and
regime blindness, are low-frequency and high-complexity. They require reasoning about
global activation geometry, the mathematical validity of candidate symbolic forms,
and the relationship between market regime and topology adaptation. These are
strategic structural decisions; a slower, larger model (200--500\,ms, 32B parameters)
is needed because these decisions require multi-step reasoning over PDE structure,
symbolic validity, and regime context.

One brain of either type cannot address all five. A single fast model makes
mathematically incoherent symbolic replacements. A single slow model blocks the GPU.
The dual-process design is not merely an efficiency tradeoff; it follows from the
structure of the problem itself. Reflexive maintenance and strategic topology surgery
are different kinds of decisions, and they should not be collapsed into a single
controller.

Our ablation study (Table~\ref{tab:ablation}) confirms the division of labor
empirically: System~2 alone achieves regime-ordered behavior but at reduced
throughput; System~1 alone preserves throughput but cannot achieve monotonicity;
only the dual-process configuration achieves both.

\subsection{Contributions}
\label{sec:intro:contributions}

\begin{enumerate}
  \item A taxonomy of five chronic management problems in heterogeneous KAN topology,
  each mapped to a specific gradient-signal gap, and a demonstration that they decompose
  across two cognitive timescales.

  \item A \textbf{dual-process topology agent} (System~1/System~2) that addresses all
  five, with the two paths operating concurrently without GPU blocking.

  \item A \textbf{typed DSL} with AST allowlist and atomic rollback providing type
  safety, vocabulary restriction ($C^2$ continuity preserved by construction), and
  transactional integrity.

  \item \textbf{Evaluation on Heston options pricing}: a testbed where $C^2$
  continuity, PDE geometry reasoning, and real-time regime adaptation impose hard
  correctness requirements that expose dual-process advantages invisible in
  unconstrained settings.

  \item The \textbf{training objective insufficiency finding}: PDE residual
  minimization alone does not produce regime-ordered behavior under distributional
  shift, a result that may extend to other physics-constrained networks under regime change.
\end{enumerate}

\subsection{Paper Organization}

Section~\ref{sec:related} situates the work in the literature.
Section~\ref{sec:background} develops the heterogeneous KAN background and the Heston
testbed. Section~\ref{sec:agent} presents the dual-process topology agent.
Section~\ref{sec:regime} describes the market regime conditioning layer.
Section~\ref{sec:experiments} presents experimental evaluation.
Section~\ref{sec:discussion} provides discussion and addresses the distinction from
LangChain-style systems. Section~\ref{sec:conclusion} concludes.

% =====================================================================
\section{Related Work}
\label{sec:related}
% =====================================================================

\subsection{KANs and Symbolic Regression}

The KAN architecture \citep{liu_etal_2024_kan} places learnable activation functions
on edges rather than nodes, enabling edge-level interpretability and surgical symbolic
replacement. The theoretical foundation is Kolmogorov's superposition theorem
\citep{kolmogorov1957representation}. MonoKAN \citep{polo_molina_etal_2024_monokan}
extends KANs with certified monotonicity constraints. Symbolic edge replacement is
available in the original KAN library but is \emph{manual and offline}. We make it
online and agent-driven, with formal invariants maintained throughout.

\subsection{Physics-Informed Learning}

PINNs \citep{raissi2019physics} convert a PDE boundary value problem into an
unconstrained optimization by embedding the differential operator residual in the
loss. Financial applications \citep{horvath2021deep} have generally used standard MLP
backbones; the $C^2$ continuity requirement of the Heston cross-derivative term has
received limited attention. Deep hedging \citep{buhler2019deep} and Neural SDEs
\citep{kidger2021neural} extend the neural approach to replication and continuous-time
dynamics, but do not enforce specific boundary conditions in the Heston sense.

\subsection{Neural Architecture Search}

NAS \citep{elsken2019neural} automates topology discovery through gradient-based
\citep{liu2019darts} or evolutionary search. NAS operates \emph{offline}: it finds
a fixed architecture before training begins. The dual-process agent operates
\emph{online}, on a running network, with natural language input and domain-specific
mathematical constraints. The two approaches are complementary; NAS for
initialization, dual-process for runtime adaptation, rather than competitive.

\subsection{Dual-Process AI and LLM Agents}

Kahneman's framework \citep{kahneman2011thinking} applied to AI agents. Toolformer
\citep{schick_etal_2023_toolformer} and ReAct \citep{yao_etal_2023_react} use LLMs
as tool orchestrators, producing text outputs over stateless API calls. The critical
distinction from these systems and from workflow frameworks (LangChain-style tool
routing) is developed in Section~\ref{sec:discussion:notlangchain}: those systems
route stateless text between API endpoints and have no persistent mathematical state
to maintain. Our agent modifies the internal mathematical structure of a running
network under formal invariants, a fundamentally different task.

% =====================================================================
\section{Background: Heterogeneous KAN Topology}
\label{sec:background}
% =====================================================================

\subsection{KAN Edge Activations}
\label{sec:background:edges}

The Kolmogorov-Arnold representation theorem \citep{kolmogorov1957representation}
guarantees that any continuous function can be expressed as a composition of
univariate functions. The KAN architecture \citep{liu_etal_2024_kan} operationalizes
this by placing learnable activation functions on \emph{edges} rather than nodes.
Each edge is implemented as a \texttt{BSplineEdge} module. The basis functions are
defined by the Cox--de Boor recursion over a knot sequence $\mathbf{t}$:
\begin{align}
B_{i,0}(x) &= \mathbf{1}[t_i \leq x < t_{i+1}] \label{eq:cox_de_boor_1} \\
B_{i,k}(x) &= \frac{x - t_i}{t_{i+k} - t_i} B_{i,k-1}(x)
             + \frac{t_{i+k+1} - x}{t_{i+k+1} - t_{i+1}} B_{i+1,k-1}(x)
\label{eq:cox_de_boor_2}
\end{align}
OpKAN uses $G=5$ interior intervals and spline order $k=3$ (cubic), giving $G+k=8$
basis functions per edge. The edge activation combines a SiLU base with a trained
spline correction:
\begin{equation}
\phi(x) = w_\text{base} \cdot \text{SiLU}(x)
         + w_\text{spline} \cdot \sum_{i=1}^{8} c_i B_{i,3}(x)
\label{eq:bspline_edge}
\end{equation}
where $w_\text{base}, w_\text{spline} \in \mathbb{R}$ and $\{c_i\}_{i=1}^8$ are
trainable. Because each $B_{i,3}$ is $C^2$ continuous (twice-differentiable with continuous second derivatives) by construction, Eq.~\eqref{eq:bspline_edge}
is $C^2$ for any choice of coefficients.

The Kolmogorov-Arnold representation theorem applies to $[0,1]^n$; in our setting the
training domain $(S,v,t) \in [0,200] \times [0,0.5] \times [0,1]$ is normalized to
$[0,1]^3$ before the collocation grid is applied. All B-spline knot sequences are
defined over this normalized domain.

The PI-KAN uses a $[3, 16, 1]$ configuration (inputs $(S,v,t)$, 16-node hidden layer,
scalar output $V$). The forward pass aggregates edge outputs by summation:
\begin{equation}
h_j^{(\ell)} = \sum_{i=1}^{n_\text{in}^{(\ell)}} \phi_{ij}^{(\ell)}\!\left(h_i^{(\ell-1)}\right)
\label{eq:kan_forward}
\end{equation}

The \emph{topology surgery problem} is formalized as follows.

\begin{definition}[Topology Surgery Problem]
Let $\mathcal{G} = (V, E, \Phi)$ be a KAN graph where $\Phi = \{\phi_e\}_{e \in E}$
assigns each edge $e$ an activation module from the parameterization class set
$\mathcal{C} = \{\texttt{BSplineEdge}, \texttt{C2SymbolicEdge}, \texttt{ZeroEdge}\}$.
Gradient-based training optimizes $\{\theta_e\}$ within each $\phi_e$'s current class
but produces no signal for reassigning $\phi_e$ to a different class in $\mathcal{C}$.
The \emph{topology surgery problem} is: given training statistics $\mathcal{S}_t$ and
domain constraints $\mathcal{K}$ (e.g., $C^2$ continuity, PDE residual geometry),
determine the reassignment $\sigma: E \to \mathcal{C}$ that improves $\mathcal{K}$-compliance
without increasing $\mathcal{L}_\text{PDE}$ beyond a tolerance $\epsilon$.
\end{definition}

\subsection{Heterogeneous Edge Types}
\label{sec:background:heterogeneous}

Three edge types coexist in OpKAN at runtime:
\begin{itemize}[nosep]
  \item \texttt{BSplineEdge}: 8 basis functions, trainable coefficients $+$ SiLU base.
  \item \texttt{C2SymbolicEdge}: closed-form expression from the 20-function AST
  allowlist, no trainable parameters.
  \item \texttt{ZeroEdge}: returns \texttt{torch.zeros\_like(x)}, implementing correct PRUNE
  semantics (not \texttt{nn.Sequential()} which is identity).
\end{itemize}

The topology at any training step is a mix of these types. This heterogeneity enables
interpretability and efficiency gains, and also creates the five chronic problems
identified in Section~\ref{sec:intro:surgery}.

\subsection{The Heston PDE Testbed}
\label{sec:background:heston}

The Heston stochastic volatility model \citep{heston1993closed} describes risk-neutral
dynamics for spot price $S_t$ and variance $v_t$:
$dS_t = r S_t \, dt + \sqrt{v_t} S_t \, dW_t^S$, $\;dv_t = \kappa(\theta - v_t) \, dt + \sigma \sqrt{v_t} \, dW_t^v$,
where $\text{Corr}(dW^S, dW^v) = \rho \, dt$. By It\^o's lemma, option fair value
$V(S,v,t)$ satisfies the Heston PDE:
\begin{equation}
\frac{\partial V}{\partial t}
+ \frac{1}{2}vS^2\frac{\partial^2 V}{\partial S^2}
+ \rho\sigma vS\frac{\partial^2 V}{\partial S\partial v}
+ \frac{1}{2}\sigma^2 v\frac{\partial^2 V}{{\partial v}^2}
+ rS\frac{\partial V}{\partial S}
+ \kappa(\theta-v)\frac{\partial V}{\partial v}
- rV = 0
\label{eq:heston_pde}
\end{equation}
with boundary conditions:
$V(S, v, T) = \max(S - K, 0)$ (terminal payoff),
$V(0, v, t) = 0$ (zero spot), and
$V(S, v, t) \approx S - Ke^{-r(T-t)}$ as $S \to \infty$.

The cross-derivative term $\rho\sigma vS \,\partial^2 V / \partial S \partial v$ is
the key constraint: evaluating it via automatic differentiation requires $C^2$
continuous activations. Standard MLP architectures with ReLU activations produce
zero second derivatives, making this term numerically degenerate.

This testbed is demanding for the dual-process agent along three axes: (1) the $C^2$
continuity requirement is a hard mathematical invariant that every topology mutation
must preserve; (2) PDE geometry reasoning constrains which symbolic functions are
mathematically coherent replacements; and (3) real-time regime adaptation requires
the agent to reason about distributional shift.

\subsection{Training Objective}
\label{sec:background:loss}

Training minimizes a composite physics-informed loss:
\begin{equation}
\mathcal{L} = \underbrace{\mathbb{E}_{\bm{x} \sim \Omega}\!\left[\left|\mathcal{N}[V](\bm{x})\right|^2\right]}_{\mathcal{L}_\text{PDE}}
+ \lambda_\text{BC}\underbrace{\mathbb{E}_{\bm{x} \sim \partial\Omega}\!\left[\left|V(\bm{x}) - V_\text{target}(\bm{x})\right|^2\right]}_{\mathcal{L}_\text{BC}}
\label{eq:loss}
\end{equation}
where $\mathcal{N}[V]$ is the Heston differential operator evaluated via
\texttt{torch.autograd} with \texttt{create\_graph=True}. Collocation points
$\bm{x}=(S,v,t)$ are drawn uniformly from $[0,200]\times[0,0.5]\times[0,1]$.
This objective is necessary but \emph{not sufficient} for regime-ordered behavior,
the key finding established in Section~\ref{sec:experiments}.

% =====================================================================
\section{Dual-Process Topology Agent}
\label{sec:agent}
% =====================================================================

\subsection{Decomposition by Cognitive Type}
\label{sec:agent:decomposition}

The five chronic problems map to two cognitive timescales as shown in
Table~\ref{tab:decomposition}. The fast path (System~1) addresses problems that can
be diagnosed from local statistics and resolved with simple actions. The slow path
(System~2) addresses problems that require multi-step mathematical reasoning over the
global network structure.

\begin{table}[ht]
\centering
\caption{Decomposition of the five chronic topology management problems by cognitive
timescale. System~1 actions are statistical threshold decisions; System~2 actions
require strategic structural reasoning over PDE geometry.}
\begin{tabular}{@{}llll@{}}
\toprule
Problem & System & Reasoning Type & Action \\
\midrule
Topology stagnation & 1 & Statistical threshold & PRUNE edge \\
$C^2$ instability & 1 & Magnitude monitoring & Adjust lr \\
Parameterization lock-in & 2 & Mathematical reasoning & REPLACE with symbolic \\
Symbolic-numerical tension & 2 & Global graph reasoning & Selective REPLACE/KEEP \\
Regime blindness & 2 & Regime analysis $+$ PDE & Structural adaptation \\
\bottomrule
\end{tabular}
\label{tab:decomposition}
\end{table}

\begin{remark}[Decomposition Sufficiency]
The decomposition in Table~\ref{tab:decomposition} is motivated by task structure, not proven
as optimal. A formal optimality result would require a decision-theoretic model of the
topology surgery problem with explicit latency-quality tradeoffs. Such a model is left
for future work. The empirical Table~\ref{tab:ablation} provides partial evidence: System~1
alone cannot achieve monotonicity (problems 2--4 require System~2's reasoning), while
System~2 alone incurs throughput cost that System~1's pruning recovers.
\end{remark}

\subsection{System 1: Reflexive Maintenance}
\label{sec:agent:sys1}

Every 10 training steps, the training loop places (step index, per-edge L1 norms,
recent loss delta) into a bounded queue (\texttt{maxsize=1}, latest-context-only).
A background thread dequeues and invokes a 7B Qwen 2.5 model, which produces a
\texttt{ReflexDecision}:

\begin{lstlisting}[language=Python]
class ReflexDecision(BaseModel):
    reasoning: str          # brief justification
    prunes: List[str]       # edge IDs with L1 norm below threshold
    lr_adjustment: float    # multiplier in [0.5, 2.0]
\end{lstlisting}

Latency must stay below 50\,ms; the 7B model is chosen because the decisions
require pattern recognition (is this norm below 0.005?) rather than deep reasoning.
This path directly addresses Problems~1 (stagnation) and~5 ($C^2$ instability) from
Section~\ref{sec:intro:surgery}: near-zero edges are pruned before they consume
further computation; learning rate adjustment prevents aggressive coefficient updates
from destabilizing second-derivative numerics.

\subsection{System 2: Strategic Surgery}
\label{sec:agent:sys2}

Every 100 training steps, a longer-horizon context (loss trajectory, HMM transition
probabilities, full KAN topology state, regime signals) is placed in a separate
bounded queue (\texttt{maxsize=1}). A second background thread invokes a 32B Qwen
2.5 model, which produces a \texttt{StrategicDecision}. System~2 is guided by PDE
geometry and market-regime signals (the regime label is produced by the HMM; see
Section~\ref{sec:regime}):

\begin{lstlisting}[language=Python]
class StrategicDecision(BaseModel):
    reasoning: str                    # strategic structural reasoning over PDE geometry
    mutations: List[EdgeMutation]     # REPLACE, PRUNE, or KEEP per edge
    regime_analysis: RegimeThesis     # HMM transition diagnosis
    training_command: Literal["CONTINUE", "HALT"]
\end{lstlisting}

The 32B model is necessary because edge replacement decisions require: (a) identifying
that an activation profile approximates a specific smooth function; (b) verifying the
candidate formula is $C^2$ on its natural domain; (c) reasoning about how the
replacement interacts with the Heston cross-derivative term. These are multi-step
mathematical reasoning tasks that 7B models perform unreliably. This path addresses
Problems~2 (parameterization lock-in), 3 (symbolic-numerical tension), and~4 (regime
blindness).

A representative System~2 output when the analyst submits \emph{``COMEX silver
inventories fell 8\% overnight; demand shock likely''}:

\begin{lstlisting}[language=Python, basicstyle=\ttfamily\footnotesize]
{
  "training_command": "CONTINUE",
  "regime_analysis": {"hmm_transition_detected": true,
                       "predicted_regime": 2,
                       "thesis_statement": "Inventory shock..."},
  "mutations": [
    {"edge_id": "L0_N0_to_L1_N3",
     "action": "REPLACE",
     "formula": "torch.exp(-torch.abs(x))",
     "reasoning": "L1 norm 0.0031 < 0.005; shape matches decaying exponential",
     "initial_params": {"scale": 1.0}}
  ]
}
\end{lstlisting}

\subsection{The Typed DSL Safety Layer}
\label{sec:agent:dsl}

Every agent output is a JSON object validated by Pydantic schema before entering the
mutation pipeline. Formula strings within \texttt{EdgeMutation.formula} are
additionally validated by static AST analysis before storage:

\begin{lstlisting}[language=Python]
_ALLOWED_TORCH_FUNCS = frozenset({
    'sin', 'cos', 'tan', 'exp', 'log', 'log1p', 'sqrt', 'abs',
    'pow', 'tanh', 'sigmoid', 'softplus', 'sign', 'clamp',
    'relu', 'floor', 'ceil', 'round', 'reciprocal', 'neg'
})
_SAFE_AST_NODES = frozenset({
    ast.Expression, ast.Call, ast.BinOp, ast.UnaryOp,
    ast.Add, ast.Sub, ast.Mult, ast.Div, ast.Pow,
    ast.FloorDiv, ast.Mod, ast.USub, ast.UAdd,
    ast.Constant, ast.Load
})

def _validate_symbolic_expression(expr: str) -> None:
    tree = ast.parse(expr, mode='eval')
    for node in ast.walk(tree):
        if type(node) in _SAFE_AST_NODES: continue
        if isinstance(node, ast.Name):
            if node.id not in ('x', 'torch'):
                raise ValueError(f"Disallowed name '{node.id}'")
        elif isinstance(node, ast.Attribute):
            if not (isinstance(node.value, ast.Name)
                    and node.value.id == 'torch'):
                raise ValueError("Only 'torch.*' attributes allowed.")
            if node.attr not in _ALLOWED_TORCH_FUNCS:
                raise ValueError(f"torch.{node.attr} not in allowlist.")
        else:
            raise ValueError(f"Disallowed node '{type(node).__name__}'")
\end{lstlisting}

\begin{proposition}
Any expression passing \texttt{\_validate\_symbolic\_expression} is restricted to
arithmetic combinations of the 20 allowlisted \texttt{torch.*} functions applied to
the variable \texttt{x}. It cannot import modules, access the file system, or invoke
arbitrary Python builtins.
\end{proposition}
\begin{proof}[Proof sketch]
The \texttt{ast.walk} traversal visits every node in the parse tree. The validator's
\texttt{else} branch (line 9) raises \texttt{ValueError} for any node type not matched
by the preceding cases. The exhaustiveness argument is: the first \texttt{if} handles
structural operators (\texttt{\_SAFE\_AST\_NODES}); the
\texttt{isinstance(node, ast.Name)} branch handles all name references; the
\texttt{isinstance(node, ast.Attribute)} branch handles all attribute access. All
remaining node types, including \texttt{ast.Subscript}, \texttt{ast.Starred},
\texttt{ast.keyword}, \texttt{ast.FormattedValue}, and others, fall into the
\texttt{else} branch and raise immediately. A formula cannot perform subscript access,
starred unpacking, or string formatting, which eliminates the main injection paths
beyond name and attribute access. No path to module import, file I/O, or builtins
survives the traversal.
\end{proof}

\begin{remark}
This is not a general-purpose sandbox. It is a domain-specific grammar restriction
that encodes the mathematical vocabulary of $C^2$ functions. The allowlist
simultaneously enforces security (no code injection) and correctness (all allowed
functions are $C^2$ on their natural domains).
\end{remark}

\begin{definition}[Semantic Compiler]
A \emph{semantic compiler} is an LLM-to-backend interface defined by the triple
$(\mathcal{S}, \mathcal{V}, \mathcal{T})$: a \emph{schema} $\mathcal{S}$ (Pydantic
model) that validates structural well-formedness; a \emph{vocabulary} $\mathcal{V}$
(AST allowlist) that restricts formula strings to a domain-specific mathematical
vocabulary; and a \emph{transaction} $\mathcal{T}$ (deepcopy snapshot + rollback)
that enforces atomic application. An LLM output passing all three components is
\emph{schema-valid}, \emph{vocabulary-safe}, and \emph{atomically applicable}.
\end{definition}

\subsection{Non-Blocking Coordination and Atomic Rollback}
\label{sec:agent:coord}

Figure~\ref{fig:architecture} shows the information flow. The \texttt{EngineCoordinator}
manages four bounded queues: \texttt{reflex\_queue} (\texttt{maxsize=1}, training
loop $\to$ System~1), \texttt{reflex\_decision\_queue} (\texttt{maxsize=32}, System~1
$\to$ training loop), \texttt{strategic\_queue} (\texttt{maxsize=1}, training loop
$\to$ System~2), and \texttt{strategic\_decision\_queue} (\texttt{maxsize=32}).

At the start of each training batch, \texttt{apply\_pending\_mutations()} non-blockingly
drains both output queues. If no decisions are pending, training continues at full GPU
throughput. The GPU never waits on the LLM.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  node distance=0.55cm and 2.2cm,
  box/.style={rectangle, rounded corners=4pt, draw=black!50, minimum width=3.6cm,
              minimum height=0.65cm, align=center, font=\small},
  databox/.style={box, fill=layerA},
  kannbox/.style={box, fill=layerB},
  agentbox/.style={box, fill=layerC},
  secbox/.style={box, fill=layerD},
  arr/.style={-Stealth, thick},
  dasharr/.style={-Stealth, dashed, thick, gray!70}
]
%% ---- Data column (left) ----
\node[databox] (market)   {Market Data (OPRA)};
\node[databox, below=of market]  (feat)    {Feature Vector $\bm{f}_t$};
\node[databox, below=of feat]    (hmm)     {HMM Regime Detector\\$r_t \in \{0,1,2\}$};

%% ---- PI-KAN (centre) ----
\node[kannbox, below=1.1cm of hmm] (pikan)
  {PI-KAN Training Loop\\(GPU: $\mathcal{L}_\text{PDE}+\lambda\mathcal{L}_\text{BC}$)};

%% ---- Agent (right of PI-KAN) ----
\node[agentbox, right=2.4cm of pikan, yshift= 0.75cm] (sys1)
  {System~1: Qwen 2.5-7B\\Reflex $<$50\,ms};
\node[agentbox, right=2.4cm of pikan, yshift=-0.75cm] (sys2)
  {System~2: Qwen 2.5-32B\\Strategy 200--500\,ms};

%% ---- Security + output ----
\node[secbox,  below=1.1cm of pikan] (mutator)
  {TopologicalMutator\\(AST-validated, atomic rollback)};
\node[kannbox, below=of mutator] (topology)
  {Updated KAN Topology\\(B-spline $\cup$ Symbolic edges)};

%% ---- Arrows ----
\draw[arr] (market)  -- (feat);
\draw[arr] (feat)    -- (hmm);
\draw[arr] (hmm)     -- (pikan);
% async bidirectional queues
\draw[dasharr] (pikan.east) to[bend left=12]
  node[above, font=\footnotesize, pos=0.45]{async queue} (sys1.west);
\draw[dasharr] (pikan.east) to[bend right=12] (sys2.west);
\draw[dasharr] (sys1.west) to[bend left=12] (pikan.east);
\draw[dasharr] (sys2.west) to[bend right=12] (pikan.east);
% downward flow
\draw[arr] (pikan)   -- (mutator);
\draw[arr] (mutator) -- (topology);
\end{tikzpicture}
\caption{OpKAN information flow across three decoupled layers. The HMM does not know
about the KAN; the topology agent does not know about the PDE internals; the GPU
training loop never waits on the LLM. Dashed arrows are asynchronous (bounded queues;
maxsize=1 for inputs, maxsize=32 for outputs). Sustained throughput of 26,300\,smp/sec
is measured during 5-epoch training; 6,506\,smp/sec is from a 50-step cold-start
microbenchmark (see Section~\ref{sec:experiments:headroom}).}
\label{fig:architecture}
\end{figure}

Atomic rollback addresses the fifth chronic problem: a failed mutation batch cannot
leave the model in partially-modified state. All affected edge modules are snapshotted
via \texttt{copy.deepcopy} before applying any mutations; any exception restores all
objects:

\begin{algorithm}
\caption{Atomic Mutation Application with Rollback}
\begin{algorithmic}[1]
\State $\textit{snapshots} \leftarrow \{\}$
\For{each edge $(l, i, j)$ in \textit{mutations}}
    \State $\textit{snapshots}[l,i,j] \leftarrow \texttt{deepcopy}(\texttt{model.layers}[l].\texttt{edges}[i][j])$
\EndFor
\State \textbf{try:}
\For{each mutation $m$ in \textit{mutations}}
    \State \texttt{TopologicalMutator.mutate\_edge}(\texttt{model}, $m$)
\EndFor
\State \textbf{except Exception as} $e$\textbf{:}
\For{each $(l,i,j) \in \textit{snapshots}$}
    \State $\texttt{model.layers}[l].\texttt{edges}[i][j] \leftarrow \textit{snapshots}[l,i,j]$
\EndFor
\State \textbf{log} rollback event with exception $e$
\end{algorithmic}
\end{algorithm}

\texttt{state\_dict} snapshots are insufficient because \texttt{swap\_edge} replaces
entire \texttt{nn.Module} objects; a \texttt{BSplineEdge} state dict cannot be loaded
into a \texttt{C2SymbolicEdge}. We snapshot module objects rather than parameter tensors.

\subsection{Current Implementation Status}
\label{sec:agent:status}

The dual-process controller as implemented is \emph{prompt-engineered}, not
\emph{policy-learned}. System~1 and System~2 execute a fixed system prompt that
instructs the LLM to output a typed DSL object; there is no gradient signal over the
LLM weights derived from downstream topology outcomes. The \texttt{LoraDataCollector}
infrastructure exists to log agent trajectories for future LoRA fine-tuning
\citep{hu2021lora}, but no fine-tuned weights have been trained. The distinction
matters: a prompt-engineered controller's behavior is contingent on prompt quality
and may not generalize to out-of-distribution inputs; a policy-learned controller
would have formal guarantees derivable from the training objective. All results in
this paper are for the prompt-engineered version. Closing the gap to a learned policy
is Limitation~9.

% =====================================================================
\section{Market Regime Conditioning}
\label{sec:regime}
% =====================================================================

\subsection{Hidden Markov Regime Detector}
\label{sec:regime:hmm}

The regime detector fits a Gaussian HMM with $K=3$ states and diagonal covariance to
five time-series features: log-return $\ell_t$, realized volatility $\hat{\sigma}_t$
(20-bar window, annualized), implied volatility $\sigma^\text{IV}_t$, IV--RV spread
$\Delta_t = \sigma^\text{IV}_t - \hat{\sigma}_t$, and IV momentum
$\mu^\text{IV}_t = \sigma^\text{IV}_t - \sigma^\text{IV}_{t-1}$.
Walk-forward inference uses non-overlapping 20-day windows; the model is refit from
scratch at each window boundary. Train/test split: the HMM is fit on the first 80\%
of the time series; regime labels for the remaining 20\% are inferred without
refitting. The label stability correction (Eq.~\ref{eq:hmm_relabel}) is applied at
every refit.

A persistent problem with rolling HMM inference is state label instability: ``state~0''
in window $t$ may map to ``state~1'' in window $t+1$ after a refit. We fix this with
a realized-volatility sorting pass:
\begin{equation}
\pi = \text{argsort}(\bm{\mu}_{:,1}), \quad \text{remap}[s] = \pi^{-1}[s]
\label{eq:hmm_relabel}
\end{equation}
where $\bm{\mu}_{:,1}$ is the vector of per-state mean realized volatilities, ensuring
label~0 always maps to the lowest-volatility regime across all rolling windows.

\subsection{Regime as Topology Signal}
\label{sec:regime:signal}

Market microstructure features are processed by the HMM to produce a regime label,
which forms part of System~2's context for topology decisions. The HMM regime label
feeds System~2's context at every strategic query. A detected regime transition
(signaled by the HMM label, not directly by analyst text) triggers a strategic request
even if fewer than 100 training steps have elapsed. Direct analyst inputs (e.g., a
text description of a supply shock) provide a separate, complementary path into
System~2's context, but they do not replace the HMM-derived regime signal. This
mechanism directly addresses Problem~4 (regime blindness). Why not let the PI-KAN
learn regime-sensitivity directly? The Heston PDE does not change with regime; it is
a structural prior on the \emph{shape} of the pricing function. What changes is the
parameters $(\kappa, \sigma, \rho)$. Conflating these two levels would undermine the
physics prior.

% =====================================================================
\section{Experimental Evaluation}
\label{sec:experiments}
% =====================================================================

\subsection{Setup}
\label{sec:experiments:setup}

\textbf{Hardware.} NVIDIA H200, CUDA. LLM inference via Ollama (local) with
\texttt{qwen2.5-coder} for protocol benchmarks; vLLM for engineering stress tests.

\textbf{Data.} SLV (silver ETF) option chains with real historical regime features.
Live benchmark: one quote date (single quote date; cross-sectional split only).
Replay benchmark: historical silver/regime feature sequence with replayed SLV surface
template.

\textbf{Model families.}
\begin{itemize}[nosep]
  \item \texttt{bs\_plain}: Black-Scholes with baseline ATM volatility.
  \item \texttt{bs\_regime}: Black-Scholes with HMM regime-predicted volatility.
  \item \texttt{rf\_direct}: Random forest direct price prediction.
  \item \texttt{spikan\_hybrid}: Symbolic PI-KAN residual $+$ regime-vol BS anchor.
  \item \texttt{spikan\_plain\_hybrid}: Symbolic PI-KAN residual $+$ plain-formula BS anchor.
  \item \texttt{spikan\_plain\_iv\_hybrid}: Plain-formula anchor $+$ learned IV-space residual (headline hybrid).
\end{itemize}

\textbf{System prompt disclosure.} The system prompt instructs the LLM to output a
JSON object conforming to the \texttt{StrategicDecision} or \texttt{ReflexDecision}
Pydantic schema, provides two few-shot examples covering each DSL operation family,
and specifies the 20-function symbolic vocabulary. The prompt structure (but not the
full text, which is domain-proprietary) is described in the supplementary materials.
Protocol benchmark results are sensitive to the few-shot examples; we report performance
only for the version with examples, leaving zero-shot ablation to future work.

\subsection{Finding 1: Protocol Validity (RQ1--RQ3)}
\label{sec:experiments:protocol}

\begin{table}[ht]
\centering
\caption{Protocol benchmark results compressed across tiers. Tier~0A: 15 curated
prompts (floor of protocol viability). Tier~0B: 8 document-grounded prompts (primary
evidence tier). Tier~0C/0E: semantic stability and gold agreement (see text).}
\begin{tabular}{@{}lccccc@{}}
\toprule
Context & $n$ & Execute & Intent & Exact DSL & Semantic \\ \midrule
Direct prompt (0A) & 9 & 1.00 & 1.00 & --- & --- \\
Narrative prompt (0A) & 6 & 1.00 & 1.00 & --- & --- \\
Document-grounded (0B) & 8 & 1.00 & 1.00 & --- & --- \\
Paraphrase stability (0C) & 4 grp & --- & --- & 0.75 (grp) & 0.9167 \\
Document-to-gold (0E) & 8 & 1.00 & --- & 0.375 & 0.75 \\ \bottomrule
\end{tabular}
\label{tab:protocol}
\end{table}

The agent correctly compiles analyst intent across all contexts (Execute=1.00 on 0A
and 0B). The semantic agreement of 0.75 on Tiers~0C and~0E (pilot study, $n \leq 15$)
reflects a normalization contract problem: the failing paraphrase group expanded a
three-point strike grid to five points (intent-preserving, structurally different),
and regime under-escalation occurred on one 0E prompt (STRESS compiled as TRANSITION).
Neither represents a comprehension failure. Both are addressable by tighter
normalization rules in the system prompt. The exact agreement ceiling of 0.375 is an
open engineering problem, not an architectural one.

These protocol results are preliminary. With $n \leq 15$ prompts per tier, all rates
should be interpreted as existence proofs of viability rather than stable probability
estimates.

\subsection{Finding 2: Training Objective Insufficiency (RQ4)}
\label{sec:experiments:monotonicity}

\begin{table}[ht]
\centering
\caption{Financial monotonicity checks under explicit DSL regime overrides (Tier~0D).
``Raw'' = model queried directly under STABLE and STRESS Heston parameter anchors.
``With Controller'' = semantic controller with monotone projection applied. All five
checks fail for the raw model; all five pass with the controller.}
\begin{tabular}{@{}lcc@{}}
\toprule
Monotonicity Check & Raw Output & With Controller \\ \midrule
Quote implied vol: STABLE $\to$ STRESS & $\times$ & \checkmark \\
Quote price: STABLE $\to$ STRESS & $\times$ & \checkmark \\
Surface mean vol: STABLE $\to$ STRESS & $\times$ & \checkmark \\
Outlook shock-score: STABLE $\to$ STRESS & $\times$ & \checkmark \\
Shock-risk labels ordered low to high & $\times$ & \checkmark \\
\bottomrule
\end{tabular}
\label{tab:coherence}
\end{table}

This is the central finding. Under the dual-process controller, all five checks pass;
without it, all five fail. This is a necessary-condition result: in our setting, the
dual-process controller is a necessary condition for regime-ordered behavior, though
we cannot yet claim it is the minimal such condition.

Both conditions (raw model, controller-assisted) use identical trained weights
evaluated on the same 500 collocation points. The difference is solely the presence
or absence of the semantic controller's regime override. This isolates the
insufficiency as a property of the training objective, not the model capacity.

PDE residual minimization is a \emph{necessary but insufficient} training objective
for regime-ordered behavior. The raw model correctly minimizes $\mathcal{L}_\text{PDE}$;
it fails all five monotonicity checks because the PDE loss imposes no constraint on
the ordering of outputs across different Heston parameter configurations. This is not
a deficiency of KANs specifically; it is a property of any physics-constrained
network trained without explicit regime-ordering terms in the objective, and may extend
to other physics-constrained networks under regime change. The
dual-process agent imposes this ordering through the semantic controller. The open
question is whether it can be embedded in the training objective through ranking losses
such as $\max(0,\, V_\text{STABLE} - V_\text{STRESS} + \epsilon)$, making monotonicity
intrinsic rather than extrinsic.

\textbf{Heston parameter anchors.} STABLE: $(\kappa{=}3.0,\,\theta{=}0.02,\,\sigma{=}0.15,\,\rho{=}{-}0.50)$.
STRESS: $(\kappa{=}0.5,\,\theta{=}0.09,\,\sigma{=}0.60,\,\rho{=}{-}0.85)$, with
$r{=}0.05$, $K{=}100$, $T{=}1$. Each model queried at 500 uniformly sampled
$(S,v,t)$ points.

\subsection{Finding 3: Pricing Performance (RQ5)}
\label{sec:experiments:pricing}

\begin{table}[ht]
\centering
\caption{Replay price MAE by market regime. Lower is better.
\texttt{spikan\_plain\_iv\_hybrid} achieves best overall MAE (0.9317), with strength
in transition and stress; the conditions where regime-conditioned corrections
matter most.}
\begin{tabular}{@{}lcccc@{}}
\toprule
Model & Stable & Transition & Stress & Overall \\ \midrule
\texttt{bs\_plain} & 0.8398 & 2.1037 & 1.6355 & 1.4597 \\
\texttt{spikan\_plain\_hybrid} & 1.1542 & 2.4827 & 0.4745 & 1.3705 \\
\texttt{spikan\_plain\_iv\_hybrid} & 1.1628 & 1.4622 & 0.4239 & \textbf{0.9317} \\
\texttt{rf\_direct} & 8.3183 & 10.2679 & 1.8707 & 6.8190 \\
\bottomrule
\end{tabular}
\label{tab:replay}
\end{table}

\begin{table}[ht]
\centering
\caption{Cross-asset results. The hybrid advantage is consistent on macro-narrative
assets (silver, oil) and absent on structural assets (semiconductors, technology),
a domain fit signal, not an architectural failure. A separate PI-KAN is trained per
asset; weights are not shared across assets. The Heston parameters $(\kappa, \theta,
\sigma, \rho)$ used for collocation are the same across assets (the canonical
STABLE/STRESS anchors), which may explain part of the weaker performance on assets
with fundamentally different volatility dynamics.}
\begin{tabular}{@{}llcc@{}}
\toprule
Asset & Dominant Signal Type & Best Model & Price MAE \\ \midrule
SLV (silver) & Macro / supply narrative & \texttt{spikan\_plain\_iv\_hybrid} & 0.9317 \\
GLD (gold) & Mean-reverting vol structure & \texttt{bs\_plain} & 1.6515 \\
USO (oil) & Inventory shock / macro & \texttt{spikan\_plain\_iv\_hybrid} & 0.4897 \\
SOXX (semiconductors) & Structural / earnings & \texttt{rf\_direct} & 1.7194 \\
XLK (technology) & Structural / earnings & \texttt{rf\_direct} & 2.5180 \\
\bottomrule
\end{tabular}
\label{tab:cross_asset}
\end{table}

The IV-residual hybrid achieves best overall replay MAE (0.9317). The cross-asset
pattern confirms the architecture's domain specificity: the hybrid wins on silver and
oil (assets where macro narrative and supply-shock interpretation are operationally
relevant), while gold, semiconductor, and technology equities favor simpler baselines.
For SOXX and XLK, weaker performance on these assets suggests that relevant structural
signals (earnings surprise, export controls, AI capex) may be absent from the current
feature set, indicating where feature engineering is needed rather than an architectural
failure.

\subsection{Finding 4: Engineering Headroom (RQ6)}
\label{sec:experiments:headroom}

\begin{table}[ht]
\centering
\caption{Engineering performance on NVIDIA H200, batch size 4,096. Sustained
throughput (26,300\,smp/sec) measured during 5-epoch training of 1M rows. The
6,506\,smp/sec E2E microbenchmark reflects cold-start amortization on 50 steps, not
a measurement error.}
\begin{tabular}{@{}lc@{}}
\toprule
Metric & Value \\ \midrule
Sustained training throughput (5 epochs, 1M rows) & 26,300\,smp/sec \\
E2E microbenchmark throughput (50 steps, cold start) & 6,506\,smp/sec \\
System~1 (fast path) latency & $<$50\,ms \\
System~2 (slow path) latency & 200--500\,ms \\
Decision queue overflows (1M-row stress test) & 0 \\
GPU blocking time from LLM & 0\,ms \\
Dual-process vs.\ baseline throughput delta & +4.4\% \\
\bottomrule
\end{tabular}
\label{tab:engineering}
\end{table}

Zero queue overflows and zero GPU blocking validate the non-blocking coordination
design. An observed 4.4\% throughput gain over a pure gradient-descent baseline
remains unattributed; no ablation isolating pruning rate, queue overhead, or batch
composition has been conducted. We treat this as a directional signal requiring future
ablation (Section~\ref{sec:limfuture}). System~1 progressive pruning of near-zero
edges is a candidate explanation for the reduced computational graph load, but it is
not the only candidate explanation. In a hypothetical synchronous design, a single
500\,ms System~2 decision would block $\approx 3{,}200$ training steps at sustained
throughput; the asynchronous design reduces this cost to zero.

\subsection{Agent Ablation}
\label{sec:experiments:ablation}

\begin{table}[ht]
\centering
\caption{Monotonicity check pass rate and training throughput by agent configuration.
No-agent baseline = pure gradient descent. Results directional; see \S\ref{sec:limfuture}
for full ablation plan.}
\begin{tabular}{@{}lcc@{}}
\toprule
Agent Configuration & Monotonicity Checks Passed & Throughput \\
\midrule
No agent (gradient only) & 0 / 5 & 25,170 smp/sec \\
System~1 only (reflexive) & 0 / 5 & 26,300 smp/sec \\
System~2 only (strategic) & 5 / 5 & 24,890 smp/sec \\
Dual-process (S1 + S2) & 5 / 5 & 26,300 smp/sec \\
\bottomrule
\end{tabular}
\label{tab:ablation}
\end{table}

The ablation reveals a clean division of labor. Monotonicity is entirely driven by
System~2's higher-latency strategic reasoning over PDE geometry and regime signals:
neither gradient-only training nor System~1 reflexive
pruning produces regime-ordered behavior (0/5 pass rate in both cases). This confirms
that monotonicity is a product of System~2's strategic structural reasoning over PDE
geometry and regime signals, not a side-effect of pruning or throughput management.
Conversely, System~2 alone incurs a throughput cost relative to the dual-process
configuration: 24,890 vs.\ 26,300\,smp/sec, a 5.3\% reduction. This cost arises
because without System~1's statistical pruning, the 32B model is invoked more
frequently on computationally expensive decisions that System~1 would have resolved
reflexively. The dual-process configuration recovers full throughput while preserving
all monotonicity guarantees, confirming that the two systems are complementary rather
than redundant.

% =====================================================================
\section{Discussion}
\label{sec:discussion}
% =====================================================================

\subsection{What the Dual-Process Architecture Solves}
\label{sec:discussion:solves}

Each of the five chronic problems maps to a finding. Problems~1 and~5 (stagnation
and $C^2$ instability) are addressed by System~1 reflexive pruning and lr control,
evidenced by zero queue overflows and sustained GPU throughput (Table~\ref{tab:engineering}).
Problems~2, 3, and~4 (parameterization lock-in, symbolic-numerical tension, regime
blindness) are addressed by System~2 strategic surgery, evidenced by all five
monotonicity checks passing only under the controller (Table~\ref{tab:coherence}).

The architecture's key insight is that the cognitive structure of the topology surgery
problem determines the design. These are not two speeds of the same task: they are
qualitatively different cognitive tasks that happen to coexist in the same system.
Forcing them into a single model creates either a bottleneck (if the slow model
handles everything) or incoherence (if the fast model attempts mathematical reasoning).

\subsection{The Training Objective Insufficiency Finding: Generalization}
\label{sec:discussion:generalization}

The monotonicity failure of the raw model may extend to other physics-constrained
networks under regime change. Any physics-constrained network
trained without regime-ordering terms will fail similar checks under distributional
shift. The PDE loss constrains the solution on the training domain; it does not
enforce behavioral ordering at out-of-distribution query points.

The semantic controller is a structural remedy. The path to making monotonicity
\emph{intrinsic} is enriching the training objective with ranking losses: penalizing
$\max(0,\, V_\text{STABLE} - V_\text{STRESS} + \epsilon)$ on sampled pairs under
the two parameter anchors. MonoKAN \citep{polo_molina_etal_2024_monokan} substitution
on the vol-surface path is a complementary architectural direction. Both remain open.

\subsection{Why This Is Not LangChain}
\label{sec:discussion:notlangchain}

LangChain-style systems route stateless text between API endpoints. Every call is
independent, every output is text, and the system has no persistent mathematical state
to maintain. The dual-process topology agent is qualitatively different along five axes:

\begin{itemize}[nosep]
  \item It manages the \textbf{continuous evolution of a heterogeneous mathematical
  object} (the KAN topology) across thousands of training steps.

  \item Its decisions must satisfy \textbf{formal invariants} ($C^2$, PDE consistency,
  monotonicity) that it reasons about explicitly, not just produce plausible text.

  \item It operates \textbf{concurrently with an ongoing computation} (GPU training
  loop) with non-blocking requirements enforced by bounded queues.

  \item Its actions are \textbf{irreversible without explicit rollback}: a topology
  mutation changes the computational graph; a \texttt{state\_dict} restore cannot
  undo a parameterization class change from \texttt{BSplineEdge} to
  \texttt{C2SymbolicEdge}.

  \item The two speeds encode \textbf{structurally different cognitive tasks}
  (statistical threshold vs.\ mathematical reasoning), not just different latencies.
\end{itemize}

The closest established category for the dual-process controller is \emph{online
structural control}: a controller that modifies the structure of a dynamical system
in real time based on observed system state and external signals, subject to invariant
constraints. This is distinct from meta-learning (which learns a policy over a
distribution of tasks offline) and from continual learning
\citep{kirkpatrick2017overcoming} (which modifies weights to avoid forgetting). The
novelty is that the ``controller'' is an LLM performing higher-latency strategic reasoning
over PDE geometry and regime signals in natural language about
mathematical structure, and the ``invariants'' are formal mathematical properties
($C^2$ continuity, typed DSL, atomic rollback).

\subsection{Limitations and Future Work}
\label{sec:limfuture}

\begin{enumerate}
  \item Single quote date for the live benchmark (generalization unestablished).
  \item Small protocol benchmark ($n \leq 15$ per tier); rates are existence proofs,
  not stable probability estimates.
  \item Exact DSL agreement ceiling of 0.375 requires system prompt engineering.
  \item No fine-tuned model weights (LoRA infrastructure exists but awaits production
  trajectory data).
  \item Monotonicity is still extrinsic to the training objective.
  \item Cross-asset advantage limited to macro-narrative domains.
  \item The 4.4\% throughput gain is directional and unablated; a full ablation
  isolating pruning rate, queue overhead, and batch composition is needed.
  \item The ablation in Table~\ref{tab:ablation} is directional; formal confidence
  intervals and larger-scale replication are needed.
  \item Closing the gap from prompt-engineered to policy-learned controller (Limitation~9).
\end{enumerate}

% =====================================================================
\section{Conclusion}
\label{sec:conclusion}
% =====================================================================

KANs' unique property, edge-level symbolic replaceability, creates a topology
surgery problem that gradient descent cannot solve. The dual-process architecture
addresses this problem by matching cognitive task type to model capability: fast
statistical pattern recognition for high-frequency reflexive maintenance (System~1,
7B, $<$50\,ms), slow mathematical reasoning for low-frequency strategic surgery
(System~2, 32B, 200--500\,ms). Demonstrated on a physics-informed KAN under Heston
PDE constraints, the architecture produces regime-ordered behavior that PDE training
alone cannot, with full GPU throughput maintained throughout.

The central finding, training objective insufficiency, may extend beyond options
pricing: any physics-constrained network trained under PDE residuals alone may fail
regime-ordering checks under distributional shift. The semantic controller is a
structural remedy; ranking losses are the path to making monotonicity intrinsic.

The central open problem: can the regime ordering enforced by the semantic controller
be embedded in the training objective through ranking losses, making monotonicity
intrinsic rather than extrinsic? That question, and the five chronic problems of
heterogeneous KAN topology, constitute the research agenda this paper opens.

% =====================================================================
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
