\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{bm}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{multirow}
\usepackage{array}
\usepackage{enumitem}
\usepackage{parskip}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric,arrows.meta,positioning,fit,backgrounds,calc}

\geometry{margin=1in}

\definecolor{codeblue}{RGB}{30,100,200}
\definecolor{layerA}{RGB}{210,230,255}  % data / regime
\definecolor{layerB}{RGB}{210,255,220}  % PI-KAN
\definecolor{layerC}{RGB}{245,220,255}  % agent
\definecolor{layerD}{RGB}{255,235,210}  % security / output
\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{codeblue}\bfseries,
  commentstyle=\color{gray},
  breaklines=true,
  frame=single
}

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

% -----------------------------------------------------------------------
\title{\textbf{Semantic Compiler: Safe LLM Orchestration of Neural Network Topology via Typed DSL}\\[0.5em]
\large\textit{A Design Pattern for Physics-Constrained Networks, Demonstrated on Real-Time Option Pricing}}
\author{
  OpKAN Engineering Team\\
  \texttt{github.com/RicardoLaMo/OpKAN}
}
\date{March 2026}

\begin{document}
\maketitle

% -----------------------------------------------------------------------
\begin{abstract}
Giving a large language model real-time structural control over a running neural
network creates a fundamental interface problem: the LLM produces unstructured text;
the network demands typed, mathematically consistent structural updates. Naive
approaches --- evaluating LLM-generated code at runtime --- introduce code injection
vulnerabilities and leave the model in partially-mutated state on failure. We present
the \textbf{semantic compiler} design pattern: the LLM's output is a typed,
Pydantic-validated domain-specific language (DSL) object; formula strings are
validated by static AST analysis before storage; and mutations are applied atomically
with \texttt{deepcopy} rollback on failure. These three properties make every
topology intervention type-safe, vocabulary-restricted, and transactionally reversible.

We demonstrate the pattern in a demanding testbed: real-time management of a
physics-informed Kolmogorov-Arnold network (PI-KAN) solving the Heston stochastic
volatility PDE, where the $C^2$ continuity requirement of the cross-derivative term
creates a hard mathematical invariant the agent must not violate. A dual-process
architecture --- 7B reflexive path ($<$50\,ms) and 32B strategic path (200--500\,ms)
--- operates concurrently with the GPU training loop with zero blocking.

Evaluation against five financial monotonicity checks reveals that PDE residual
training alone is \emph{insufficient} for regime-ordered behavior --- monotonicity
requires the semantic controller --- identifying training objective insufficiency as a
general failure mode in physics-constrained networks under distributional shift.
Protocol benchmarks achieve 1.00 execution rate and 0.75 semantic agreement against
human-authored gold DSL, with the residual gap identifying a normalization contract
problem rather than a comprehension failure. Sustained throughput reaches
$\mathbf{26{,}300}$\,smp/sec on NVIDIA H200 with zero GPU blocking.
\end{abstract}

\tableofcontents
\newpage

% =====================================================================
\section{Introduction}
\label{sec:intro}
% =====================================================================

\subsection{The Interface Problem: LLM-Driven Topology Management}

Consider a neural network solving a physics PDE that is currently training on a GPU.
An analyst reads a market narrative and wants the network's topology updated to reflect
the new regime --- specific edges replaced with symbolic functions, others pruned ---
within seconds, without stopping training. This is a topology management problem, and
giving an LLM control over it creates a specific interface challenge that has no
satisfactory solution in existing literature.

The mismatch is structural. The LLM produces natural language or loosely-structured
JSON; the network accepts only typed, mathematically valid structural mutations. Three
requirements must be simultaneously satisfied: (1) \textbf{Type safety}: the mutation
must conform to a well-defined schema before it reaches the network; (2)
\textbf{Vocabulary restriction}: any formula proposed must be limited to a safe
mathematical vocabulary that preserves the network's invariants (in our case, $C^2$
continuity for Heston PDE evaluation); (3) \textbf{Transactional atomicity}: if a
mutation batch fails mid-execution, the model must be restored to its pre-mutation
state.

Naive approaches fail all three simultaneously. Direct evaluation of LLM-generated
Python via \texttt{eval()} violates type safety (no schema enforcement), vocabulary
restriction (any Python builtin is accessible), and atomicity (partial mutations
persist on failure). Neural architecture search \citep{elsken2019neural} solves none
of these problems: it operates offline before training, receives no natural language
input, and provides no mechanism for enforcing domain-specific mathematical invariants.
Prompt-to-price approaches --- where the LLM produces a numerical price directly ---
are not topology management at all: they bypass the learned model entirely and
introduce hallucination risk on numerical outputs.

\subsection{The Semantic Compiler Approach}

The semantic compiler resolves this by positioning the LLM as a compiler: it takes
unstructured natural language and produces structured, typed output. The
``compilation'' step enforces three properties before the output ever reaches the
network:

\begin{itemize}[nosep]
  \item \textbf{Pydantic schema validation at parse time.} The LLM is instructed to
  produce JSON conforming to a typed schema; any output that fails validation triggers
  an error-correction retry rather than downstream execution.

  \item \textbf{AST allowlist analysis on formula strings.} Any formula string within
  a command is checked by static analysis against a pre-approved mathematical vocabulary
  before it is stored. The vocabulary is domain-specific: it encodes the set of
  functions that are $C^2$ continuous on their natural domains, blocking all other
  names including module imports and file-system access.

  \item \textbf{deepcopy snapshot $+$ rollback for transactional integrity.} All
  affected modules are snapshotted before mutation begins. Any exception during
  application restores all objects. Either all mutations in a batch succeed or none do.
\end{itemize}

The LLM never produces a price, never modifies weights directly, and never executes
code. It produces a DSL object. The specialist backend handles execution.

\subsection{The Testbed: Why Heston/KAN Is a Demanding Domain}

The finance domain is not the primary contribution of this paper, but it is chosen
as a testbed because it stress-tests the semantic compiler pattern in three ways that
synthetic benchmarks cannot:

\begin{itemize}[nosep]
  \item \textbf{$C^2$ continuity constraint.} The agent must propose formulas that are
  twice-differentiable on the physical domain, not merely syntactically valid Python.
  This makes the vocabulary layer's mathematical content non-trivial: the allowlist
  encodes a genuine mathematical property, not just a security policy.

  \item \textbf{PDE geometry reasoning.} Strategic topology decisions require
  understanding which edge activations are consistent with the Heston differential
  operator's structure --- a reasoning task that exercises the 32B strategic model's
  domain comprehension at each decision step.

  \item \textbf{Real-time latency.} A System-2 decision cannot block GPU training,
  motivating the dual-process architecture. The asynchronous design is not a
  performance optimization; it is a correctness requirement for any online topology
  management system.
\end{itemize}

\subsection{Contributions}

This paper makes the following contributions:

\begin{enumerate}
  \item \textbf{The semantic compiler design pattern} (Typed DSL $+$ AST allowlist
  $+$ atomic rollback) as a reusable approach for safe LLM-driven topology management
  of physics-constrained networks.

  \item \textbf{A dual-process non-blocking architecture} (System~1 reflexive, System~2
  strategic) with a formal zero-blocking guarantee enforced by bounded queues and
  non-blocking drain logic.

  \item \textbf{PI-KAN for Heston}: a complete implementation with $C^2$ B-splines
  via Cox--de Boor recursion, verified by second-order autograd on the Heston
  cross-derivative term.

  \item \textbf{A tiered evaluation framework} spanning protocol validity, paraphrase
  stability, normalization gap, regime coherence, pricing performance, and engineering
  headroom --- designed to independently assess each layer of the architecture.

  \item \textbf{The training objective insufficiency finding}: PDE residual
  minimization alone does not produce regime-ordered behavior. Monotonicity requires
  the semantic controller, identifying a general failure mode in physics-constrained
  networks under distributional shift.
\end{enumerate}

\subsection{Scope and Positioning}
\label{sec:scope}

OpKAN is not a pricing formula and does not claim to outperform all existing pricing
models in all conditions. Its specific claim is that the semantic compiler pattern
makes LLM-driven topology management safe, auditable, and non-blocking in a domain
where violations of mathematical invariants have immediate operational consequences.
The cross-asset experiments in Section~\ref{sec:results} show precisely where the
architecture's advantages hold and where domain coverage limits them.

\subsection{Paper Organization}

Section~\ref{sec:related} situates the work in the literature.
Section~\ref{sec:formulation} develops the problem formulation.
Section~\ref{sec:pattern} defines the semantic compiler pattern in the abstract.
Section~\ref{sec:arch} describes the OpKAN instantiation.
Section~\ref{sec:security} presents correctness guarantees.
Section~\ref{sec:experiments} presents the experimental design.
Section~\ref{sec:results} reports findings. Section~\ref{sec:discussion} interprets them.
Section~\ref{sec:limfuture} states scope boundaries and outlines future directions.
Section~\ref{sec:conclusion} concludes.

% =====================================================================
\section{Related Work}
\label{sec:related}
% =====================================================================

OpKAN sits at the intersection of six research areas. We review each with emphasis on
the specific gaps that motivated our design choices.

\subsection{Physics-Informed Learning for Financial PDEs}

PINNs \citep{raissi2019physics} convert a PDE boundary value problem into an
unconstrained optimization by embedding the differential operator residual in the loss
function. Applications to financial PDEs \citep{horvath2021deep} have generally used
standard MLP backbones. The $C^2$ continuity requirement of the Heston cross-derivative
term has received limited attention; in practice, many implementations approximate or
omit the $\rho\sigma vS$ coefficient entirely. OpKAN's PI-KAN backbone addresses this
gap by using activations whose continuity class is guaranteed by construction.

Finite difference methods \citep{wilmott1993option} and Monte Carlo simulation remain
standard Heston baselines. They are accurate but rigid: re-solving under shifted
parameters requires re-running the full solver. Deep hedging \citep{buhler2019deep}
extends the neural approach to replication and hedging under real-world dynamics;
Neural SDEs \citep{kidger2021neural} provide continuous-time neural network models
that are related architecturally to our PDE-constrained training but do not enforce
specific boundary conditions in the Heston sense.

\subsection{Neural Architecture Search and Topology Management}

Neural architecture search (NAS) \citep{elsken2019neural} automates the discovery of
optimal network topologies through gradient-based \citep{liu2019darts} or evolutionary
search. The semantic compiler approach is architecturally distinct along three axes.
First, NAS operates \emph{offline}: it finds a fixed architecture before training begins,
whereas the topology agent performs \emph{online} surgery on a network that is actively
training. Second, NAS receives no natural language input; the architecture space is
explored algorithmically with no semantic grounding. Third, NAS does not enforce
domain-specific mathematical invariants during search: there is no mechanism to require
that discovered activations be $C^2$ continuous, or that a proposed edge replacement be
consistent with PDE geometry.

Continual and lifelong learning \citep{kirkpatrick2017overcoming} addresses a related
problem --- how to modify a network without catastrophic forgetting --- but does not
address the interface between natural language and topology mutation. The semantic
compiler does not manage forgetting; it manages the safety and correctness of the
mutation itself.

\subsection{Kolmogorov-Arnold Networks}

The KAN architecture \citep{liu_etal_2024_kan} places learnable activation functions on
edges rather than nodes, enabling edge-level interpretability and surgical symbolic
replacement. The theoretical foundation is Kolmogorov's superposition theorem
\citep{kolmogorov1957representation}. The key practical advantage over MLPs is the
combination of $C^2$ B-spline activations and the ability to replace individual edges
with symbolic functions.

MonoKAN \citep{polo_molina_etal_2024_monokan} extends KANs with certified monotonicity
constraints --- the direction OpKAN's architecture should evolve. The current system
achieves regime-ordered pricing through an external semantic controller, not native
architectural monotonicity; selective MonoKAN substitution is a stated future direction
(Section~\ref{sec:limfuture}).

\subsection{LLM Tool Use and Financial NLP}

Toolformer \citep{schick_etal_2023_toolformer} and ReAct \citep{yao_etal_2023_react}
establish that LLMs can learn to invoke external tools and APIs. Domain-specific
financial LLMs such as FinBERT \citep{huang2023finbert} and BloombergGPT
\citep{wu2023bloomberggpt} demonstrate that pre-training on financial text improves
structured information extraction.

OpKAN's topology management agent takes a different position from all of these: the LLM
is a \emph{source of structured commands}, not a direct decision-maker or a
named-entity extractor. Its output is a typed DSL object validated by a Pydantic schema
before it is applied, and the validation failure itself is auditable. This distinction
has operational consequences: a generic tool call can fail silently or produce
financially inconsistent outputs; a DSL command cannot be applied if it fails schema
validation.

\subsection{Safe and Sandboxed Code Execution}

The problem of safely executing LLM-generated code has been approached through runtime
sandboxes (restricted interpreter environments, OS-level isolation) and static analysis
(type checking, taint tracking). Runtime sandboxes impose significant overhead and
cannot enforce domain-specific constraints such as $C^2$ continuity. Static type
checking operates at a coarser granularity than expression-level allowlisting.

The AST allowlist approach in Section~\ref{sec:security} occupies a different point in
this design space: it is a \emph{compile-time domain grammar restriction} that doubles
as a mathematical vocabulary constraint. Rather than asking ``is this code safe to
execute?'' (the sandbox question), it asks ``is this expression within the mathematical
vocabulary I have pre-approved for $C^2$ network activations?'' The vocabulary
constraint is tighter than any general-purpose sandbox while being lighter-weight and
directly enforceable at parse time.

\subsection{Market Regime Detection}

Gaussian HMMs for market regime detection follow Hamilton's foundational work
\citep{hamilton1989new} on nonlinear regime-switching time series. The label stability
problem under rolling re-estimation is well known but rarely addressed explicitly in
applied work. Our realized-volatility sorting approach (Section~\ref{sec:arch:hmm}) is
grounded in the empirical regularity that realized volatility is the most discriminative
feature for separating market regimes in our feature set.

\subsection{Dual-Process AI Architectures}

Kahneman's dual-process framework \citep{kahneman2011thinking} has been applied to AI
agent design in robotics and game-playing. The specific combination of an LLM fast path
and an LLM slow path operating on the same bounded queues --- with both paths producing
typed, schema-validated outputs that feed the same mutation pipeline --- is, to our
knowledge, a novel application to neural network topology management.

% =====================================================================
\section{Problem Formulation}
\label{sec:formulation}
% =====================================================================

Before describing the architecture, we formalize the constraints that any solution must
satisfy. This section establishes three requirements that jointly motivate the design
choices in Section~\ref{sec:arch}.

\subsection{The Heston PDE and the $C^2$ Constraint}

The Heston stochastic volatility model \citep{heston1993closed} describes risk-neutral
asset dynamics as a joint process for spot price $S_t$ and instantaneous variance $v_t$:
\begin{align}
dS_t &= r S_t \, dt + \sqrt{v_t} S_t \, dW_t^S \\
dv_t &= \kappa(\theta - v_t) \, dt + \sigma \sqrt{v_t} \, dW_t^v
\end{align}
where $\kappa$ is mean-reversion speed, $\theta$ is the long-run variance, $\sigma$ is
volatility-of-volatility, and $\text{Corr}(dW^S, dW^v) = \rho \, dt$ captures the
leverage effect. By It\^o's lemma, the fair value $V(S,v,t)$ of a European call
satisfies the Heston PDE:
\begin{equation}
\frac{\partial V}{\partial t}
+ \frac{1}{2}vS^2\frac{\partial^2 V}{\partial S^2}
+ \rho\sigma vS\frac{\partial^2 V}{\partial S\partial v}
+ \frac{1}{2}\sigma^2 v\frac{\partial^2 V}{{\partial v}^2}
+ rS\frac{\partial V}{\partial S}
+ \kappa(\theta-v)\frac{\partial V}{\partial v}
- rV = 0
\label{eq:heston_pde}
\end{equation}
supplemented by boundary conditions:
\begin{align}
V(S, v, T) &= \max(S - K, 0) \quad \text{(terminal payoff)} \label{eq:bc1} \\
V(0, v, t) &= 0 \quad \text{(zero spot)} \label{eq:bc2} \\
V(S, v, t) &\approx S - Ke^{-r(T-t)} \quad \text{as } S \to \infty \label{eq:bc3}
\end{align}

The cross-derivative term $\rho\sigma vS \,\partial^2 V / \partial S \partial v$ is the
key constraint for architecture design. Evaluating this term via automatic
differentiation requires the network activation functions to be \emph{at least $C^2$
continuous}: once-differentiable activations (ReLU, leaky ReLU) produce zero second
derivatives and make this term numerically degenerate, ruling out standard MLP
architectures for faithful Heston PINN training.

\subsection{The Regime Problem: Why a Single Learned Model Is Insufficient}

A single trained neural network, even a well-constrained PINN, learns a \emph{fixed}
mapping from inputs $(S,v,t)$ to prices. The Heston parameters $(\kappa,\theta,\sigma,
\rho)$ are not constants --- they shift with market regime. A single frozen model trained
on historical data averaged across regimes will systematically underprice
out-of-the-money options in stress conditions and overprice them in calm conditions.
The regime problem is a consequence of the non-stationarity of the underlying stochastic
process, not a defect of any particular architecture.

\subsection{The Narrative Problem: Why Structured Models Cannot Self-Update}

Even with a regime-sensitive architecture, a purely quantitative system has no channel
for the most time-sensitive market information: qualitative narrative. Central bank
announcements, geopolitical developments, and supply-shock narratives arrive as natural
language. Converting them to structured inputs requires either a human analyst (high
latency, not scalable) or a system that reliably translates unstructured text into
structured quantitative commands.

This is not primarily a machine learning problem --- it is an interface design problem.
The challenge is to define an interface that accepts free-form language as input and
produces structured, validated quantitative commands as output, so that the downstream
pricing model never receives unvalidated, potentially inconsistent inputs.

These three requirements --- $C^2$ continuous activations, regime-adaptive topology, and a
structured interface for narrative inputs --- jointly motivate the semantic compiler
pattern in Section~\ref{sec:pattern} and its OpKAN instantiation in Section~\ref{sec:arch}.

% =====================================================================
\section{The Semantic Compiler Pattern}
\label{sec:pattern}
% =====================================================================

Before describing the OpKAN instantiation, we abstract the design pattern it
implements. The pattern applies to any system where an LLM must modify the structure
of a running constraint-satisfying computational process. The Heston PI-KAN is one
instance; the pattern is reusable wherever similar interface requirements arise.

\subsection{The Core Problem: Type Safety at the LLM-Network Boundary}

The fundamental mismatch is representational. An LLM's natural output space is
unstructured text or loosely-structured JSON; a physics-constrained neural network's
mutation interface requires typed, validated commands. Three failure modes occur when
this boundary is not enforced:

\begin{enumerate}[nosep]
  \item \textbf{Schema failure.} The LLM produces a structurally invalid command
  (missing required field, wrong type). Without schema validation, this reaches the
  mutation pipeline and causes a runtime error at an unpredictable depth.

  \item \textbf{Vocabulary failure.} The LLM proposes a formula string containing
  operations outside the safe mathematical vocabulary (e.g., importing a module,
  accessing the file system, calling a disallowed function). Without static analysis,
  this executes with the full privileges of the training process.

  \item \textbf{Atomicity failure.} A mutation batch applies successfully to the
  first three edges and fails on the fourth. Without rollback, the model is left in
  a partially-mutated state with no record of which edges were modified.
\end{enumerate}

The semantic compiler resolves all three.

\subsection{Pattern Definition}

\begin{definition}[Semantic Compiler]
A \emph{semantic compiler} is an LLM-to-backend interface with three enforcement layers:
\begin{enumerate}[nosep]
  \item \textbf{Type layer.} A Pydantic schema defines the grammar of valid commands.
  The LLM is instructed to produce JSON conforming to this schema; any output that
  fails schema validation triggers an error-correction retry rather than downstream
  execution.

  \item \textbf{Vocabulary layer.} For any formula string within a command, a static
  AST analysis pass verifies membership in a pre-approved vocabulary before the
  formula is stored. The vocabulary is domain-specific: in OpKAN it encodes the set
  of functions that are $C^2$ continuous on their natural domains.

  \item \textbf{Transaction layer.} All affected backend objects are snapshotted via
  \texttt{deepcopy} before mutation begins. Any exception during application restores
  all objects and logs the failure. Either all mutations in a batch succeed or none do.
\end{enumerate}
\end{definition}

Together, these three layers provide a formal guarantee: any command that reaches the
backend is \emph{well-typed} (passes Pydantic), \emph{vocabulary-safe} (passes AST
analysis), and \emph{applied atomically} (transactional rollback). The LLM never
produces a price, never modifies weights directly, and never executes code; it produces
a DSL object. All execution happens in a typed, validated, reversible backend.

\subsection{Why This Over the Alternatives}

\begin{description}[nosep]
  \item[Direct LLM output (price or weight update).]
  The LLM produces a numerical price or a weight delta directly. This removes the
  interface entirely but eliminates auditability, introduces hallucination risk on
  numerical outputs, and cannot represent structured topology operations such as
  edge pruning or symbolic replacement.

  \item[Prompt-to-eval() (LLM generates Python, eval runs it).]
  The LLM generates arbitrary Python that is evaluated at runtime. This is a code
  injection vulnerability. A prompt-injected formula string
  \texttt{\_\_import\_\_("os").system("rm -rf /")} executes with the full privileges
  of the training process. Runtime sandboxes add overhead without enforcing
  domain-specific mathematical vocabulary.

  \item[Neural architecture search.]
  NAS discovers optimal topology offline through gradient or evolutionary search.
  It does not accept natural language as input, does not operate on a running
  network, and does not enforce domain constraints during search. NAS and the
  semantic compiler are complementary, not alternatives: NAS could be used to
  initialize the topology; the semantic compiler manages it at runtime.
\end{description}

\subsection{Generalizability}

The semantic compiler pattern is not finance-specific. Any domain where an LLM must
modify a running constraint-satisfying system --- a chemical synthesis simulator
where reaction pathways must satisfy thermodynamic constraints, a physical simulation
where mesh refinements must preserve conservation laws, a robotics controller where
trajectory modifications must remain within kinematic limits --- presents the same
three-layer interface problem. The OpKAN instantiation demonstrates viability in a
domain where the constraints are mathematically sharp and the failure modes are
operationally consequential.

% =====================================================================
\section{Architecture}
\label{sec:arch}
% =====================================================================

\subsection{Design Philosophy: One Problem per Layer}

The OpKAN architecture is organized around a single principle: each layer should do
exactly one thing well, and the boundaries between layers should be explicit, typed, and
enforceable.

\begin{itemize}[nosep]
  \item \textbf{Layer 1 (PI-KAN):} Solve the Heston PDE. Know nothing about market
  regimes or LLM outputs.
  \item \textbf{Layer 2 (HMM):} Produce a stable regime label from market
  microstructure features. Know nothing about network topology or language.
  \item \textbf{Layer 3 (Topology Agent):} Translate regime signals and topology
  statistics into typed mutation commands. Know nothing about the PDE internals or
  the raw market data.
\end{itemize}

Figure~\ref{fig:architecture} shows the information flow. Data moves in one direction
(market data $\to$ features $\to$ regime label $\to$ updated network), with the agent
operating asynchronously so it never blocks the GPU training loop.

% ------------------------------------------------------------------
\begin{figure}[t]
\centering
\begin{tikzpicture}[
  node distance=0.55cm and 2.2cm,
  box/.style={rectangle, rounded corners=4pt, draw=black!50, minimum width=3.6cm,
              minimum height=0.65cm, align=center, font=\small},
  databox/.style={box, fill=layerA},
  kannbox/.style={box, fill=layerB},
  agentbox/.style={box, fill=layerC},
  secbox/.style={box, fill=layerD},
  arr/.style={-Stealth, thick},
  dasharr/.style={-Stealth, dashed, thick, gray!70}
]
%% ---- Data column (left) ----
\node[databox] (market)   {Market Data (OPRA)};
\node[databox, below=of market]  (feat)    {Feature Vector $\bm{f}_t$};
\node[databox, below=of feat]    (hmm)     {HMM Regime Detector\\$r_t \in \{0,1,2\}$};

%% ---- PI-KAN (centre) ----
\node[kannbox, below=1.1cm of hmm] (pikan)
  {PI-KAN Training Loop\\(GPU: $\mathcal{L}_\text{PDE}+\lambda\mathcal{L}_\text{BC}$)};

%% ---- Agent (right of PI-KAN) ----
\node[agentbox, right=2.4cm of pikan, yshift= 0.75cm] (sys1)
  {System~1: Qwen 2.5-7B\\Reflex $<$50\,ms};
\node[agentbox, right=2.4cm of pikan, yshift=-0.75cm] (sys2)
  {System~2: Qwen 2.5-32B\\Strategy 200--500\,ms};

%% ---- Security + output ----
\node[secbox,  below=1.1cm of pikan] (mutator)
  {TopologicalMutator\\(AST-validated, atomic rollback)};
\node[kannbox, below=of mutator] (topology)
  {Updated KAN Topology\\(B-spline $\cup$ Symbolic edges)};

%% ---- Arrows ----
\draw[arr] (market)  -- (feat);
\draw[arr] (feat)    -- (hmm);
\draw[arr] (hmm)     -- (pikan);
% async bidirectional queues
\draw[dasharr] (pikan.east) to[bend left=12]
  node[above, font=\footnotesize, pos=0.45]{async queue} (sys1.west);
\draw[dasharr] (pikan.east) to[bend right=12] (sys2.west);
\draw[dasharr] (sys1.west) to[bend left=12] (pikan.east);
\draw[dasharr] (sys2.west) to[bend right=12] (pikan.east);
% downward flow
\draw[arr] (pikan)   -- (mutator);
\draw[arr] (mutator) -- (topology);
\end{tikzpicture}
\caption{OpKAN information flow across three decoupled layers. The HMM does not know
about the KAN; the topology agent does not know about the PDE; the GPU training loop
never waits on the LLM. Dashed arrows are asynchronous (bounded queues; maxsize=1 for
inputs, maxsize=32 for outputs). The full training throughput of 26,300\,smp/sec is
measured during sustained 5-epoch training; the standalone E2E validation suite reports
6,506\,smp/sec under a 50-step cold-start microbenchmark (see Section~\ref{sec:rq6}).}
\label{fig:architecture}
\end{figure}
% ------------------------------------------------------------------

\subsection{Layer 1: The Physics-Informed KAN}
\label{sec:arch:pikan}

\subsubsection{Why KAN Over MLP for Heston}

The Kolmogorov-Arnold representation theorem \citep{kolmogorov1957representation}
guarantees that any continuous function $f:[0,1]^n\to\mathbb{R}$ can be expressed as a
composition of univariate continuous functions:
\begin{equation}
f(x_1, \ldots, x_n) = \sum_{q=0}^{2n} \Phi_q\!\left(\sum_{p=1}^{n} \phi_{q,p}(x_p)\right)
\label{eq:ka}
\end{equation}
The domain in Eq.~\eqref{eq:ka} is the unit hypercube $[0,1]^n$; in practice the
training domain $(S,v,t)\in[0,200]\times[0,0.5]\times[0,1]$ is bounded and the
collocation grid is normalized accordingly. The KAN architecture \citep{liu_etal_2024_kan}
operationalizes this by placing learnable activation functions on \emph{edges} rather
than nodes. This has two concrete consequences for our problem.

First, edge activations can be parameterized as B-splines, which are \emph{genuinely
$C^2$ continuous}, satisfying the Heston cross-derivative requirement that standard MLPs
cannot meet.

Second, individual edges can be replaced with closed-form symbolic expressions without
modifying the rest of the network --- the topological handle the agent uses to inject
domain knowledge. The resulting network, where some edges are B-splines and some are
closed-form functions, is referred to as a \textbf{Symbolic PI-KAN}.

\subsubsection{B-Spline Edge Activations}

Each edge is implemented as a \texttt{BSplineEdge} module. The basis functions are
defined by the Cox--de Boor recursion over a knot sequence $\mathbf{t}$:
\begin{align}
B_{i,0}(x) &= \mathbf{1}[t_i \leq x < t_{i+1}] \\
B_{i,k}(x) &= \frac{x - t_i}{t_{i+k} - t_i} B_{i,k-1}(x)
             + \frac{t_{i+k+1} - x}{t_{i+k+1} - t_{i+1}} B_{i+1,k-1}(x)
\label{eq:cox_de_boor}
\end{align}
OpKAN uses $G=5$ interior intervals and spline order $k=3$ (cubic), giving $G+k=8$
basis functions per edge. The edge activation combines a SiLU base with a trained
spline correction:
\begin{equation}
\phi(x) = w_\text{base} \cdot \text{SiLU}(x)
         + w_\text{spline} \cdot \sum_{i=1}^{8} c_i B_{i,3}(x)
\label{eq:bspline_edge}
\end{equation}
where $w_\text{base}, w_\text{spline} \in \mathbb{R}$ and $\{c_i\}_{i=1}^8$ are
trainable. Because each $B_{i,3}$ is $C^2$ by construction, Eq.~\eqref{eq:bspline_edge}
is $C^2$ for any choice of coefficients.

An edge with near-zero spline coefficients can be \emph{pruned} (replaced by a
\texttt{ZeroEdge} that returns \texttt{torch.zeros\_like(x)}); an edge whose activation
profile approximates a smooth function can be \emph{replaced} by a
\texttt{C2SymbolicEdge}.

\subsubsection{Model Architecture and Training Objective}

The PI-KAN uses a $[3, 16, 1]$ configuration (three inputs $(S,v,t)$, 16-node hidden
layer, scalar output $V$). The forward pass aggregates edge outputs by summation:
\begin{equation}
h_j^{(\ell)} = \sum_{i=1}^{n_\text{in}^{(\ell)}} \phi_{ij}^{(\ell)}\!\left(h_i^{(\ell-1)}\right)
\label{eq:kan_forward}
\end{equation}

Training minimizes a composite physics-informed loss over randomly sampled collocation
points:
\begin{equation}
\mathcal{L} = \underbrace{\mathbb{E}_{\bm{x} \sim \Omega}\!\left[\left|\mathcal{N}[V](\bm{x})\right|^2\right]}_{\mathcal{L}_\text{PDE}}
+ \lambda_\text{BC}\underbrace{\mathbb{E}_{\bm{x} \sim \partial\Omega}\!\left[\left|V(\bm{x}) - V_\text{target}(\bm{x})\right|^2\right]}_{\mathcal{L}_\text{BC}}
\label{eq:loss}
\end{equation}
where $\mathcal{N}[V]$ is the Heston differential operator evaluated via
\texttt{torch.autograd} with \texttt{create\_graph=True} for second-order computation.
Interior collocation points $\bm{x} = (S,v,t)$ are drawn uniformly from
$[0,200]\times[0,0.5]\times[0,1]$ each batch. A critical implementation note: these
points are sampled randomly and \emph{independently of observed market data}. Using
market-observed $(S,v,t)$ triples as collocation points would concentrate the PDE
residual in historically observed regions, biasing the model away from the full domain.
This look-ahead bias was present in earlier versions of this codebase and is corrected
here; a quantitative before/after ablation is deferred to future work.

\subsection{Layer 2: Hidden Markov Regime Detector}
\label{sec:arch:hmm}

\subsubsection{Why a Separate Regime Layer}

Why not let the PI-KAN learn regime-sensitivity directly from data? The Heston PDE does
not change with regime --- it is a structural prior on the \emph{shape} of the pricing
function. What changes with regime is the \emph{parameters} $\kappa$, $\sigma$, $\rho$.
Conflating these two levels of adaptation would force the PI-KAN to simultaneously
learn a PDE solution and a parameter estimation procedure, undermining the physics prior.
A separate regime detector provides a clean conditioning signal.

\subsubsection{Features and Walk-Forward Inference}

The regime detector fits a Gaussian HMM with diagonal covariance to five time-series
features:
\begin{itemize}[nosep]
  \item $\ell_t = \log(S_t / S_{t-1})$ (log-return)
  \item $\hat{\sigma}_t = \text{std}(\ell_{t-W:t}) \cdot \sqrt{252 \times 6.5 \times 60}$ (realized vol, $W=20$)
  \item $\sigma^\text{IV}_t$ (implied volatility from the options chain)
  \item $\Delta_t = \sigma^\text{IV}_t - \hat{\sigma}_t$ (IV--RV spread)
  \item $\mu^\text{IV}_t = \sigma^\text{IV}_t - \sigma^\text{IV}_{t-1}$ (IV momentum)
\end{itemize}

A persistent problem with rolling HMM inference is state label instability: ``state 0''
in window $t$ may map to ``state 1'' in window $t+1$ after a refit. We fix this with a
realized-volatility sorting pass:
\begin{equation}
\pi = \text{argsort}(\bm{\mu}_{:,1}), \quad \text{remap}[s] = \pi^{-1}[s]
\label{eq:hmm_relabel}
\end{equation}
where $\bm{\mu}_{:,1}$ is the vector of per-state mean realized volatilities, ensuring
that label 0 always maps to the lowest-volatility regime across all rolling windows.

\subsection{Layer 3: The Topology Management Agent}
\label{sec:arch:agent}

\subsubsection{The Core Idea: LLM as Semantic Compiler}

The topology management agent, internally named the \emph{LiuClaw} agent, addresses a
fundamental interface problem. On one side is unstructured text; on the other is a
quantitative backend that accepts only typed mutation commands (prune this edge, replace
that activation with this formula). The agent's job is to translate one representation
into the other --- making it, in the precise sense of the term, a \emph{semantic
compiler}.

The crucial design constraint is type-safety at parse time. The LLM is instructed to
generate JSON conforming to the \texttt{StrategicDecision} or \texttt{ReflexDecision}
Pydantic schema; the schema validates the output before it enters the mutation pipeline.
A malformed response triggers an error-correction retry rather than a downstream
failure.

\subsubsection{Two-Speed Architecture}

The agent operates on two timescales, motivated by the qualitatively different tasks
each handles.

\medskip
\noindent\textbf{Fast path (System~1 --- Reflexive Maintenance).}
Every 10 training steps, the training loop places the step index, per-edge L1 activation
norms, and the recent loss delta into a bounded queue. A background thread dequeues
this, passes it to a 7B-parameter model (Qwen 2.5-7B), and writes back a
\texttt{ReflexDecision}:

\begin{lstlisting}[language=Python]
class ReflexDecision(BaseModel):
    reasoning: str      # brief justification
    prunes: List[str]   # edge IDs with near-zero L1 norm
    lr_adjustment: float  # learning-rate multiplier in [0.5, 2.0]
\end{lstlisting}

This path prunes low-norm edges and adjusts the learning rate. Latency must stay below
50\,ms, which is why the smallest available model is used.

\medskip
\noindent\textbf{Slow path (System~2 --- Strategic Surgery).}
Every 100 training steps, a longer-horizon context --- loss trajectory, HMM transition
probabilities, full KAN state --- is placed in a separate bounded queue. A second
background thread passes it to a 32B-parameter model (Qwen 2.5-32B) and writes back a
\texttt{StrategicDecision}:

\begin{lstlisting}[language=Python]
class StrategicDecision(BaseModel):
    reasoning: str                    # chain-of-thought over PDE geometry
    mutations: List[EdgeMutation]     # REPLACE, PRUNE, or KEEP per edge
    regime_analysis: RegimeThesis     # HMM transition diagnosis
    training_command: Literal["CONTINUE", "HALT"]
\end{lstlisting}

This path replaces B-spline edges with symbolic functions that the agent reasons are
consistent with the Heston PDE geometry. The 32B model is used because edge
replacement requires coherent mathematical reasoning across multiple PDE constraints.

This bifurcation implements Kahneman's dual-process model \citep{kahneman2011thinking}:
System~1 (fast, cheap, pattern recognition) never blocks System~2 (slow, deliberate,
multi-step reasoning).

\subsubsection{The Domain-Specific Language}

Every agent output is a JSON object conforming to a Pydantic schema. The atomic unit
is the \texttt{EdgeMutation}:
\begin{itemize}[nosep]
  \item \texttt{edge\_id}: form \texttt{L\{layer\}\_N\{in\}\_to\_L\{layer+1\}\_N\{out\}}
  \item \texttt{action}: one of \texttt{PRUNE}, \texttt{REPLACE}, \texttt{KEEP}
  \item \texttt{formula}: for \texttt{REPLACE}, a PyTorch expression from the allowlisted vocabulary
  \item \texttt{initial\_params}: optional \texttt{\{scale, shift, coeff\}} for warm initialization
  \item \texttt{reasoning}: quantitative justification citing L1 norms and sampled activation values
\end{itemize}

To make the interface concrete, the following is a representative end-to-end trace:
an analyst submits the text \emph{``COMEX silver inventories fell 8\% overnight; demand
shock likely''}, and the agent emits:

\begin{lstlisting}[language=Python,basicstyle=\ttfamily\footnotesize]
{
  "training_command": "CONTINUE",
  "regime_analysis": {"hmm_transition_detected": true,
                       "predicted_regime": 2,
                       "thesis_statement": "Inventory shock..."},
  "mutations": [
    {"edge_id": "L0_N0_to_L1_N3",
     "action": "REPLACE",
     "formula": "torch.exp(-torch.abs(x))",
     "reasoning": "L1 norm 0.0031 < 0.005; shape matches decaying exponential",
     "initial_params": {"scale": 1.0}}
  ]
}
\end{lstlisting}

The schema validates this JSON object (types, Literal constraints, field presence)
before it reaches the mutation pipeline; it is rejected and retried if any field
violates the schema.

\subsection{End-to-End Coordination}
\label{sec:arch:coord}

The \texttt{EngineCoordinator} manages four bounded queues:
\begin{itemize}[nosep]
  \item \texttt{reflex\_queue} (maxsize=1): training loop $\to$ System~1 thread.
  Latest-context-only: old context discarded if not yet consumed.
  \item \texttt{reflex\_decision\_queue} (maxsize=32): System~1 thread $\to$ training loop.
  \item \texttt{strategic\_queue} (maxsize=1): training loop $\to$ System~2 thread.
  \item \texttt{strategic\_decision\_queue} (maxsize=32): System~2 thread $\to$ training loop.
\end{itemize}

At the start of each training batch, \texttt{apply\_pending\_mutations(model, optimizer)}
non-blockingly drains both output queues. If no decisions are pending, training
continues at full GPU throughput. The GPU never waits on the LLM.

% =====================================================================
\section{Correctness Guarantees}
\label{sec:security}
% =====================================================================

The topology agent introduces an attack surface that conventional neural networks do not
have: it evaluates LLM-generated Python expressions inside a running training process.

\subsection{The Code Injection Problem}

The original implementation evaluated formula strings via Python's \texttt{eval()}:
\begin{lstlisting}[language=Python]
out = eval(self.expression_str, {"torch": torch}, {"x": x})
\end{lstlisting}
This is trivially exploitable. A formula string
\texttt{'\ldots\_\_import\_\_("os").system("rm -rf /")'} executes a shell command at
evaluation time with the full privileges of the training process. Since the formula
string originates from an LLM --- which could be adversarially prompted or subject to
prompt injection from document inputs --- this is a critical vulnerability in any
production deployment.

\subsection{AST Allowlist Validation}

We fix this with a static AST analysis pass that runs \emph{before the expression is
stored}:

\begin{lstlisting}[language=Python]
_ALLOWED_TORCH_FUNCS = frozenset({
    'sin', 'cos', 'tan', 'exp', 'log', 'log1p', 'sqrt', 'abs',
    'pow', 'tanh', 'sigmoid', 'softplus', 'sign', 'clamp',
    'relu', 'floor', 'ceil', 'round', 'reciprocal', 'neg'
})
_SAFE_AST_NODES = frozenset({
    ast.Expression, ast.Call, ast.BinOp, ast.UnaryOp,
    ast.Add, ast.Sub, ast.Mult, ast.Div, ast.Pow,
    ast.FloorDiv, ast.Mod, ast.USub, ast.UAdd,
    ast.Constant, ast.Load
})

def _validate_symbolic_expression(expr: str) -> None:
    tree = ast.parse(expr, mode='eval')
    for node in ast.walk(tree):
        if type(node) in _SAFE_AST_NODES: continue
        if isinstance(node, ast.Name):
            if node.id not in ('x', 'torch'):
                raise ValueError(f"Disallowed name '{node.id}'")
        elif isinstance(node, ast.Attribute):
            if not (isinstance(node.value, ast.Name)
                    and node.value.id == 'torch'):
                raise ValueError("Only 'torch.*' attributes allowed.")
            if node.attr not in _ALLOWED_TORCH_FUNCS:
                raise ValueError(f"torch.{node.attr} not in allowlist.")
        else:
            raise ValueError(f"Disallowed node '{type(node).__name__}'")
\end{lstlisting}

\begin{proposition}
Any expression passing \texttt{\_validate\_symbolic\_expression} is restricted to
arithmetic combinations of the 20 allowlisted \texttt{torch.*} functions applied to
the variable \texttt{x}. It cannot import modules, access the file system, or invoke
arbitrary Python builtins.
\end{proposition}
\begin{proof}[Proof sketch]
The \texttt{ast.walk} traversal visits every node in the parse tree. The validator
partitions all possible AST node types into three exhaustive cases and raises on any
unexpected case: \emph{(i)} safe structural nodes (\texttt{\_SAFE\_AST\_NODES}) are
passed through; \emph{(ii)} \texttt{ast.Name} nodes are permitted only for identifiers
\texttt{x} and \texttt{torch}, blocking all other variable references and built-in
names (including \texttt{\_\_import\_\_}, \texttt{eval}, \texttt{open}); \emph{(iii)}
\texttt{ast.Attribute} nodes are permitted only as \texttt{torch.\{attr\}} where
\texttt{attr} is in the 20-function allowlist, blocking chained attribute access
(\texttt{\_\_class\_\_.\_\_bases\_\_}). Any node type not in these three cases raises
immediately. No path to module import, file I/O, or builtins survives the traversal.
\end{proof}

\begin{remark}
The allowlist vocabulary was designed around $C^2$ continuity requirements: all 20
functions are smooth on their natural domains. A future hardened version should
additionally verify differentiability at runtime by probing second derivatives on a
sample grid.
\end{remark}

\subsection{Correct Pruning Semantics}

An earlier implementation represented a pruned edge as \texttt{nn.Sequential()}, the
identity under default PyTorch, making \texttt{PRUNE} semantically equivalent to
\texttt{KEEP}. We replace this with:

\begin{lstlisting}[language=Python]
class ZeroEdge(nn.Module):
    def forward(self, x: torch.Tensor) -> torch.Tensor:
        return torch.zeros_like(x)
\end{lstlisting}

\texttt{ZeroEdge} correctly removes the edge's contribution from downstream aggregation,
enabling genuine network sparsification.

\subsection{Atomic Mutation Rollback}

A mutation batch may fail mid-execution if a formula passes AST validation but fails at
runtime. Without rollback, the model is left in a \emph{partially mutated} state. We
prevent this by snapshotting all affected edge modules via \texttt{copy.deepcopy}
before applying any mutations and restoring them on any exception:

\begin{algorithm}
\caption{Atomic Mutation Application with Rollback}
\begin{algorithmic}[1]
\State $\textit{snapshots} \leftarrow \{\}$
\For{each edge $(l, i, j)$ in \textit{mutations}}
    \State $\textit{snapshots}[l,i,j] \leftarrow \texttt{deepcopy}(\texttt{model.layers}[l].\texttt{edges}[i][j])$
\EndFor
\State \textbf{try:}
\For{each mutation $m$ in \textit{mutations}}
    \State \texttt{TopologicalMutator.mutate\_edge}(\texttt{model}, $m$)
\EndFor
\State \textbf{except Exception as} $e$\textbf{:}
\For{each $(l,i,j) \in \textit{snapshots}$}
    \State $\texttt{model.layers}[l].\texttt{edges}[i][j] \leftarrow \textit{snapshots}[l,i,j]$
\EndFor
\State \textbf{log} rollback event with exception $e$
\end{algorithmic}
\end{algorithm}

\texttt{state\_dict} snapshots are insufficient because \texttt{swap\_edge} replaces
entire \texttt{nn.Module} objects; a \texttt{BSplineEdge} state dict cannot be loaded
into a \texttt{C2SymbolicEdge}. We snapshot module objects rather than parameter tensors.

% =====================================================================
\section{Experimental Investigation}
\label{sec:experiments}
% =====================================================================

\subsection{Research Questions}

Our experiments are preliminary investigations into six questions targeting specific
layers of the architecture. We frame them as questions rather than hypotheses to signal
that the current results raise follow-up problems rather than resolving them.

\begin{enumerate}[label=\textbf{RQ\arabic*}]
  \item \textbf{Protocol validity.} Can the agent reliably compile direct and narrative
  prompts into valid, executable DSL while preserving the analyst's domain intent?
  Does performance degrade gracefully or catastrophically with longer document context?

  \item \textbf{Paraphrase stability.} Is the compiled DSL stable under paraphrase?
  Does rewording the same analyst intent produce a functionally equivalent DSL command?

  \item \textbf{Normalization gap.} How large is the gap between the agent producing a
  \emph{valid} DSL command and producing the \emph{same} DSL command a human analyst
  would have produced?

  \item \textbf{Regime coherence.} Does routing a fixed query through STABLE, TRANSITION,
  and STRESS regime contexts produce directionally ordered backend outputs? This tests
  whether the explicit semantic controller works correctly and whether PDE residual
  minimization alone is sufficient for regime ordering.

  \item \textbf{Pricing performance.} Which combination of physics model and learned
  residual produces the lowest price MAE in each regime? Is the hybrid advantage
  consistent across assets?

  \item \textbf{Engineering headroom.} Does the asynchronous dual-queue architecture
  maintain full GPU training throughput under concurrent LLM interaction?
\end{enumerate}

\subsection{Experimental Setup}

\textbf{Hardware.} All GPU experiments run on NVIDIA H200 with CUDA. LLM inference uses
Ollama (local) with \texttt{qwen2.5-coder} for protocol benchmarks and vLLM for the
engineering stress test. We use a code-specialized model for the protocol benchmarks
because the DSL has strongly code-like syntax (typed JSON schema, identifier
conventions), and code models demonstrate superior structured-output adherence in our
preliminary tests.

\textbf{Data.} The pricing benchmarks use SLV (silver ETF) option chains with real
historical regime features. The live benchmark uses one quote date (cross-sectional
split). The replay benchmark uses a historical silver/regime feature sequence with a
replayed SLV surface template, enabling evaluation across stable, transition, and
stress dates.

\textbf{Model families.} We compare six families. The \texttt{spikan\_*} variants all
use a Symbolic PI-KAN (B-spline edges with some agent-replaced symbolic edges) as the
base model; they differ in their pricing formula anchor and residual type:
\begin{itemize}[nosep]
  \item \texttt{bs\_plain}: Black-Scholes with baseline ATM volatility.
  \item \texttt{bs\_regime}: Black-Scholes with HMM regime-predicted volatility.
  \item \texttt{rf\_direct}: Random forest direct price prediction from structured features.
  \item \texttt{spikan\_hybrid}: Symbolic PI-KAN residual added to a \emph{regime-vol} Black-Scholes anchor.
  \item \texttt{spikan\_plain\_hybrid}: Symbolic PI-KAN residual added to a \emph{plain-formula} Black-Scholes anchor (no regime conditioning on the anchor).
  \item \texttt{spikan\_plain\_iv\_hybrid}: Plain-formula anchor with a learned \emph{IV-space} residual. This is the headline hybrid: it corrects the implied volatility surface rather than the price directly.
\end{itemize}

\textbf{Evaluation metrics.} Protocol tiers: execution rate, intent-match rate, DSL
validity rate, semantic signature match, group consistency. Pricing tiers: price MAE by
regime, IV MAE on the live surface. Engineering: throughput (smp/sec), queue overflow
count, GPU blocking time (ms).

\textbf{Note on sample sizes.} The protocol benchmarks involve 15--23 prompts and 4
paraphrase groups. At these sizes, individual results (especially group-level pass rates
such as 3/4 = 0.75) should be interpreted as directional estimates rather than precise
measurements. We do not report confidence intervals because bootstrapping 23 binary
outcomes produces very wide intervals that would obscure rather than illuminate the
findings. The appropriate response to this limitation is to increase benchmark scale
in future work, not to over-interpret the current point estimates.

\subsection{Protocol Benchmark Tiers}

\textbf{Tier 0A (Curated benchmark):} 15 prompts (9 direct, 6 narrative) spanning
\texttt{OUTLOOK}, \texttt{QUOTE}, \texttt{SURFACE}, and \texttt{SCENARIO} operations,
written in the style of the system prompt and few-shot examples. This tier establishes
the \emph{floor of protocol viability} --- it does not measure generalization to
out-of-distribution prompts.

\textbf{Tier 0B (Document-grounded benchmark):} 8 prompts derived from real local
document excerpts (fluid dynamics notes, COMEX inventory summaries, regime-shift
research notes) with analyst requests appended. This tier tests robustness to longer,
noisier context and is the primary evidence tier for deployment planning.

\textbf{Tier 0C (Semantic stability benchmark):} Paraphrase groups, each containing
multiple phrasings of the same analyst intent. Tests whether the DSL action is
invariant to surface variation.

\textbf{Tier 0D (Regime-action coherence benchmark):} Holds the query family fixed and
varies only the DSL regime instruction (\texttt{STABLE}, \texttt{TRANSITION},
\texttt{STRESS}). The five financial monotonicity checks test whether the backend
responds with directionally ordered outputs under each regime context.

\textbf{Tier 0E (Document-to-gold agreement):} Human-authored gold DSL commands are
specified in advance for each document-grounded prompt. Agent output is compared at
three levels: exact normalized string equality, semantic signature equality, and
executed-result equality. This is the strictest tier and the primary measure of
distance from full analyst automation.

% =====================================================================
\section{Results and Findings}
\label{sec:results}
% =====================================================================

\subsection{RQ1 -- Protocol Validity}

\begin{table}[ht]
\centering
\caption{Protocol results by context tier. Execute = fraction producing valid,
executable DSL. Intent = fraction matching the analyst's stated goal. Tier~0A is a
floor check (prompts written in-distribution with the system prompt); Tier~0B is the
primary evidence tier (derived from real document excerpts).}
\begin{tabular}{@{}lcccc@{}}
\toprule
Context & Cases & Execute & Intent & Mean latency (s) \\ \midrule
Direct prompt (0A) & 9 & 1.00 & 1.00 & 0.776 \\
Narrative prompt (0A) & 6 & 1.00 & 1.00 & 0.573 \\
Document-grounded (0B) & 8 & 1.00 & 1.00 & 2.747 \\
\textbf{Overall (0A)} & \textbf{15} & \textbf{1.00} & \textbf{1.00} & \textbf{0.695} \\
\bottomrule
\end{tabular}
\label{tab:protocol}
\end{table}

All 15 curated and all 8 document-grounded prompts produce valid, executable DSL. The
Tier~0A result (1.00 on in-distribution prompts) establishes viability but not
generalization. The Tier~0B result --- maintaining 1.00 intent-match on prompts derived
from documents the system prompt was not designed around --- is the more meaningful
finding.

The $4\times$ latency increase from direct (0.78\,s) to document-grounded (2.75\,s)
reflects a genuine deployment tradeoff: document-grounded latency is acceptable for
the slow strategic path (System~2, 100-step period) but not for the fast reflexive
path (System~1, must stay under 50\,ms).

Representative compiled outputs from Tier~0B:
\begin{itemize}[nosep]
  \item Fluid-dynamics excerpt $\to$ \texttt{OUTLOOK asset=silver horizon=10d backdrop=STRESS} (shock-risk score 13.45)
  \item COMEX inventory note $\to$ \texttt{QUOTE call spot=31 strike=32 expiry=45d vol=kan\_regime backdrop=STRESS rate=5\%}
  \item Regime-shift challenge $\to$ \texttt{SURFACE vol asset=silver regime=STRESS strikes=[0.9,1.0,1.1] T=30d}
\end{itemize}

\subsection{RQ2 -- Paraphrase Stability}

\begin{table}[ht]
\centering
\caption{Semantic stability benchmark (Tier~0C). ``Group'' metrics require all
paraphrases within a group to produce the same DSL action and execution result.
$n=4$ paraphrase groups; interpret the 0.75 group pass rate as a directional estimate
from a small sample.}
\begin{tabular}{@{}lc@{}}
\toprule
Metric & Value \\ \midrule
Case translation success & 1.00 \\
Case valid DSL & 1.00 \\
Case execution success & 1.00 \\
Case semantic signature match & 0.9167 \\
Group signature consistency & 0.75 \\
Group result consistency & 0.75 \\
All-group pass rate & 0.75 \\ \bottomrule
\end{tabular}
\label{tab:stability}
\end{table}

Three of four paraphrase groups fully preserve both the DSL action and the executed
outcome across all phrasings. The failing group (\texttt{transition\_surface}) exposes a
specific normalization ambiguity: one paraphrase caused the agent to expand a
three-point strike grid to five points while correctly identifying the regime and
operation. This is a normalization failure (different structural choice) rather than a
comprehension failure (wrong intent).

The 0.75 group pass rate --- noting the small $n=4$ caveat above --- suggests that one in
four equivalent analyst queries may produce a behaviorally different but intent-preserving
command. Closing this gap is a system-prompt engineering problem (Section~\ref{sec:limfuture}), not an architectural one.

\subsection{RQ3 -- Normalization Gap}

\begin{table}[ht]
\centering
\caption{Document-to-gold agreement (Tier~0E; $n=8$ prompts). ``Exact'' = normalized
string equality with the human-authored gold DSL. ``Semantic'' = same DSL operation
family and regime. ``Result'' = same executed numerical output.}
\begin{tabular}{@{}lc@{}}
\toprule
Metric & Value \\ \midrule
Translation success & 1.00 \\
Exact DSL agreement & 0.375 \\
Valid DSL & 1.00 \\
Semantic agreement & 0.75 \\
Execution success & 1.00 \\
Result agreement & 0.75 \\ \bottomrule
\end{tabular}
\label{tab:gold}
\end{table}

The gap between exact agreement (0.375) and semantic/result agreement (0.75) is the
central diagnostic signal of the evaluation. The system reliably produces valid,
semantically aligned DSL but does not reliably produce the \emph{exact} form a human
analyst would have chosen. Both failure modes are interpretable: \emph{regime
under-escalation} (STRESS compiled as TRANSITION) and \emph{surface grid broadening}
(three-point grid expanded to five). Neither represents a semantic misunderstanding;
both represent different but valid structural choices. The path to improvement is
contract tightening on the system prompt, not architectural change.

\subsection{RQ4 -- Regime Coherence}
\label{sec:rq4}

To test regime coherence, we define two Heston parameter sets that represent canonical
market conditions. The \textbf{STABLE} anchor uses $(\kappa{=}3.0,\,\theta{=}0.02,\,
\sigma{=}0.15,\,\rho{=}{-}0.50)$, representing a calm, mean-reverting market with low
vol-of-vol. The \textbf{STRESS} anchor uses $(\kappa{=}0.5,\,\theta{=}0.09,\,
\sigma{=}0.60,\,\rho{=}{-}0.85)$, representing a crash-risk environment with slow
mean-reversion and high vol-of-vol. Both sets use $r{=}0.05$, $K{=}100$, $T{=}1$. We
query each model at 500 uniformly sampled $(S,v,t)$ points and check five monotonicity
conditions.

\begin{table}[ht]
\centering
\caption{Financial monotonicity checks under explicit DSL regime overrides (Tier~0D).
``Raw'' = model queried directly under the STABLE/STRESS Heston parameter anchors.
``With Semantic Controller'' = explicit controller with monotone projection applied
over raw outputs. All five checks fail for the raw model; all five pass with the
controller.}
\begin{tabular}{@{}lcc@{}}
\toprule
Monotonicity Check & Raw Output & With Semantic Controller \\ \midrule
Quote implied vol: STABLE $\to$ STRESS & $\times$ & \checkmark \\
Quote price: STABLE $\to$ STRESS & $\times$ & \checkmark \\
Surface mean vol: STABLE $\to$ STRESS & $\times$ & \checkmark \\
Outlook shock-score: STABLE $\to$ STRESS & $\times$ & \checkmark \\
Shock-risk labels ordered low to high & $\times$ & \checkmark \\
\bottomrule
\end{tabular}
\label{tab:coherence}
\end{table}

This result requires careful interpretation. The raw model failing all five checks does
not mean the model is wrong; it means PDE residual minimization is an
\emph{insufficient} training objective for producing regime-ordered behavior when
queried at out-of-distribution parameter sets. The model correctly minimizes what it
was trained on. The key architectural insight is that \emph{monotonicity is currently
a property of the control layer, not of the learned weights}. Whether monotonicity
should be built into the training objective (via ranking losses) or the architecture
(via MonoKAN substitution) is the most direct open research question this result raises.

\subsection{RQ5 -- Pricing Performance}

\begin{table}[ht]
\centering
\caption{Replay price MAE by market regime. Lower is better. \texttt{spikan\_plain\_iv\_hybrid}
adds a learned IV-space residual to a plain Black-Scholes anchor (no regime conditioning
on the anchor); \texttt{spikan\_plain\_hybrid} adds a price-space residual to the same
anchor. See Section~\ref{sec:experiments} for full model definitions.}
\begin{tabular}{@{}lcccc@{}}
\toprule
Model & Stable & Transition & Stress & Overall \\ \midrule
\texttt{bs\_plain} & 0.8398 & 2.1037 & 1.6355 & 1.4597 \\
\texttt{spikan\_plain\_hybrid} & 1.1542 & 2.4827 & 0.4745 & 1.3705 \\
\texttt{spikan\_plain\_iv\_hybrid} & 1.1628 & 1.4622 & 0.4239 & \textbf{0.9317} \\
\texttt{rf\_direct} & 8.3183 & 10.2679 & 1.8707 & 6.8190 \\
\bottomrule
\end{tabular}
\label{tab:replay}
\end{table}

\begin{table}[ht]
\centering
\caption{Live term-surface results (one quote date, cross-sectional split; see
Limitation~1). IV MAE measures the ability to reconstruct the observed implied
volatility surface.}
\begin{tabular}{@{}lcc@{}}
\toprule
Model & Price MAE & IV MAE \\ \midrule
\texttt{spikan\_plain\_hybrid} & 0.8976 & 0.1578 \\
\texttt{rf\_direct} & 1.0444 & \textbf{0.0781} \\
\texttt{bs\_plain} & 1.4646 & 0.1601 \\
\texttt{spikan\_hybrid} & 1.6793 & 0.1863 \\
\texttt{bs\_regime} & 12.6503 & 0.8314 \\ \bottomrule
\end{tabular}
\label{tab:live}
\end{table}

The IV-residual hybrid achieves the best overall replay MAE (0.9317), with particular
strength in transition (1.46) and stress (0.42) --- the conditions where regime-conditioned
corrections matter most. In stable conditions, plain Black-Scholes (0.84) is the best
single model, consistent with the formula being most accurate when market conditions
match its assumptions. The random-forest baseline is the best IV reconstructor (IV
MAE 0.0781) but the worst replay pricer overall (6.82) --- the canonical pattern of a
model that minimizes in-distribution reconstruction error but generalizes poorly under
distributional shift.

\begin{table}[ht]
\centering
\caption{Cross-asset pricing summary. The best model family and its price MAE vary
substantially across assets, reflecting different regime signal relevance. See text
for interpretation.}
\begin{tabular}{@{}llcc@{}}
\toprule
Asset & Dominant Signal Type & Best Model & Price MAE \\ \midrule
SLV (silver) & Macro / supply narrative & \texttt{spikan\_plain\_iv\_hybrid} & 0.9317 \\
GLD (gold) & Mean-reverting vol structure & \texttt{bs\_plain} & 1.6515 \\
USO (oil) & Inventory shock / macro & \texttt{spikan\_plain\_iv\_hybrid} & 0.4897 \\
SOXX (semiconductors) & Structural / earnings & \texttt{rf\_direct} & 1.7194 \\
XLK (technology) & Structural / earnings & \texttt{rf\_direct} & 2.5180 \\
\bottomrule
\end{tabular}
\label{tab:cross_asset}
\end{table}

The cross-asset results confirm the architecture's domain specificity: the hybrid wins
on silver and oil (assets where macro narrative and supply-shock interpretation are
operationally relevant), while gold, semiconductor, and technology equities favor
simpler baselines. For gold, the mean-reverting vol structure is well-captured by the
Black-Scholes formula. For SOXX and XLK, the relevant pricing signals (earnings
surprise, export controls, AI capital expenditure) are absent from the current feature
set. This is a signal about where feature engineering investment is needed, not a
failure of the architecture.

\subsection{RQ6 -- Engineering Performance}
\label{sec:rq6}

\begin{table}[ht]
\centering
\caption{Engineering performance on NVIDIA H200, batch size 4,096. Training throughput
(26,300\,smp/sec) is measured during sustained 5-epoch training of 1M rows; the
standalone \texttt{validate\_e2e.py} microbenchmark reports 6,506\,smp/sec on a 50-step
cold-start run with a lighter synthetic dataset. Both configurations use identical
hardware and batch size; the difference reflects startup amortization, not a
measurement error.}
\begin{tabular}{@{}lc@{}}
\toprule
Metric & Value \\ \midrule
Sustained training throughput (5 epochs, 1M rows) & 26,300\,smp/sec \\
E2E microbenchmark throughput (50 steps, cold start) & 6,506\,smp/sec \\
System~1 (fast path) latency & $<$50\,ms \\
System~2 (slow path) latency & 200--500\,ms \\
Decision queue overflows (1M-row stress test) & 0 \\
GPU blocking time from LLM & 0\,ms \\
Dual-process vs.\ baseline throughput delta & +4.4\% \\
\bottomrule
\end{tabular}
\label{tab:engineering}
\end{table}

The asynchronous dual-queue architecture maintains full GPU throughput with zero queue
overflows and zero GPU blocking. The dual-process configuration shows +4.4\% higher
effective throughput than a pure gradient-descent baseline in the 1M-row stress test.
A likely contributor is System~1's progressive pruning of near-zero edges, which
reduces the computational graph size over time. However, we do not have an ablation
isolating this effect (e.g., pruning count vs. throughput trajectory), so this remains
a hypothesis for future investigation rather than a confirmed causal claim.

In a hypothetical synchronous design, a 500\,ms System~2 decision would block
$\approx 3{,}200$ training steps at sustained throughput. The asynchronous design
reduces this cost to zero.

% =====================================================================
\section{Discussion}
\label{sec:discussion}
% =====================================================================

\subsection{What the Architecture Gets Right}

Three claims receive preliminary support.

\textbf{The semantic compiler interface is viable on Tier~0B inputs.} All 8
document-grounded prompts (derived from documents not used in constructing the system
prompt) produce valid, executable, intent-matching DSL. This is the meaningful evidence
threshold: it shows the protocol layer survives context not of its own design.

\textbf{Separation of concerns makes the system debuggable.} Because each layer is
evaluated independently, failures are localizable: a protocol failure does not implicate
the PDE solver; a monotonicity failure does not implicate the LLM compiler. This
debuggability is a structural property of the architecture, not an empirical result.

\textbf{The IV-residual hybrid is the right pricing architecture for the target use
case.} The regime-by-regime MAE profile (best in transition and stress, weaker in
stable) is exactly the pattern the architecture predicts: the learned correction adds
value when conditions deviate from the analytic formula's assumptions. This is a
structural argument, not just a post-hoc observation.

\subsection{The Monotonicity Gap: A Training Objective Problem}

The most important finding is that PDE residual minimization is not a sufficient
training objective for regime-ordered pricing. The model is not wrong; it minimizes
what it was trained on. The architectural question this raises is whether the training
objective can be enriched with regime-ordering terms --- e.g., ranking losses on pairs
of STABLE/STRESS outputs --- such that monotonicity becomes intrinsic rather than
requiring an external controller. This is the primary open design question.

\subsection{The Normalization Gap: A Contract Problem}

The gap between exact DSL agreement (0.375) and semantic agreement (0.75) on Tier~0E
is structurally similar to a software calling-convention problem: the compiler is
correct, but the contract is underspecified. Both failure modes (regime under-escalation
and grid broadening) can be addressed by tighter normalization rules in the system
prompt without architectural change.

\subsection{Cross-Asset Transfer: What Domain Specificity Means}

The cross-asset heterogeneity is not a failure of the architecture --- it is a signal
about domain fit. The architecture provides value where narrative regime interpretation
is part of the pricing problem (silver, oil). It provides no value where the pricing
signal is structural and not in the current feature set (semiconductor earnings,
technology capex). This tells us precisely which feature engineering is needed to
extend the architecture's advantage to equity derivatives.

% =====================================================================
\section{Limitations and Future Directions}
\label{sec:limfuture}
% =====================================================================

We state the following as definite scope boundaries, each paired with the specific
next step that would address it.

\begin{enumerate}
  \item \textbf{Single quote date for the live benchmark.} The live term-surface benchmark
  uses one quote date with a cross-sectional split --- a sanity check, not a
  generalization result. \emph{Next step}: obtain multi-date historical OPRA tapes
  (or the public CBOE historical options dataset) and repeat the replay benchmark
  over a rolling 12-month window.

  \item \textbf{Structural replay, not real historical options.} The replay benchmark
  uses a replayed SLV surface template, not real historical quotes. Stress-regime MAE
  values should be interpreted as directional evidence, not production-ready estimates.
  \emph{Next step}: acquire a real multi-date historical options chain for the same
  replay period.

  \item \textbf{Curated and near-distribution protocol benchmarks.} The 15 Tier~0A
  prompts were written in-distribution with the system prompt; they establish viability,
  not adversarial robustness. Tier~0B provides stronger evidence but is still curated.
  \emph{Next step}: construct a benchmark drawn from raw financial news feeds with
  adversarial phrasings.

  \item \textbf{Small benchmark sample sizes.} With $n \leq 15$ prompts per tier,
  group-level pass rates (e.g., 3/4 = 0.75) are estimated from very few data points.
  \emph{Next step}: scale the benchmark to $\geq 100$ prompts per tier before drawing
  deployment conclusions.

  \item \textbf{Exact DSL agreement ceiling of 0.375.} The remaining normalization
  failures on Tier~0E are interpretable but real. \emph{Next step}: add a monotone
  escalation rule and canonical grid specification to the system prompt; run ablations
  on Tier~0E to establish the ceiling achievable without fine-tuning.

  \item \textbf{Monotonicity is extrinsic.} All five monotonicity checks pass because
  of the semantic controller, not the raw model. \emph{Next step (a)}: add
  regime-ordering ranking losses to the training objective --- sample pairs $(S,v,t)$
  under STABLE and STRESS parameter sets and penalize $\max(0,\, V_\text{STABLE} -
  V_\text{STRESS} + \epsilon)$. \emph{Next step (b)}: selectively replace B-spline
  edges with MonoKAN-style certified monotone variants \citep{polo_molina_etal_2024_monokan}
  on the vol-surface path where the controller correction is heaviest.

  \item \textbf{Domain-limited cross-asset advantage.} The hybrid does not universally
  outperform simpler baselines. \emph{Next step}: for SOXX and XLK, add earnings
  surprise indicators, export-control event flags, and AI capital expenditure indices
  as HMM input features, then repeat the cross-asset experiments.

  \item \textbf{Security analysis is proof-of-concept.} The AST allowlist validator
  blocks all five injection payloads tested but has not been subjected to formal
  red-teaming. \emph{Next step}: engage a security review against a broader adversarial
  payload suite; additionally add runtime differentiability probing on a sample grid.

  \item \textbf{No fine-tuned model weights.} The \texttt{LoraDataCollector}
  infrastructure exists for collecting LiuClaw agent trajectories. \emph{Next step}:
  once sufficient production trajectory data has been collected, apply LoRA
  \citep{hu2021lora} fine-tuning to the 7B model (System~1 reflexive task) and the
  32B model (System~2 strategic task) to reduce latency and close the normalization gap.

  \item \textbf{Retrieval and adversarial defense for the document input path.} Moving
  from curated excerpts to a live news feed requires a retrieval pipeline and input
  sanitization separate from the AST execution-side validator. \emph{Next step}:
  design a document ingestion layer with deduplication and retrieval-level injection
  defense.
\end{enumerate}

% =====================================================================
\section{Conclusion}
\label{sec:conclusion}
% =====================================================================

We have presented the semantic compiler as a design pattern for safe, auditable
LLM-driven topology management of physics-constrained neural networks, and demonstrated
it in the OpKAN system --- a three-layer architecture for real-time options pricing
under the Heston stochastic volatility model.

The pattern's three enforcement layers --- Pydantic type validation, AST allowlist
analysis, and atomic mutation rollback --- together provide a boundary at which every
LLM-driven network modification is type-safe, vocabulary-restricted, and
transactionally reversible. The dual-process agent (System~1 reflexive, System~2
strategic) ensures neither LLM path blocks the GPU training loop.

The most consequential finding is not about pricing accuracy: it is that PDE residual
minimization is \emph{insufficient} as a training objective for regime-ordered
behavior. The raw model correctly minimizes what it was trained on; regime ordering
requires the semantic controller. This finding generalizes beyond options pricing to
any physics-constrained network deployed under distributional shift: the physics prior
constrains the solution on the training domain, but does not enforce behavioral
ordering at out-of-distribution query points. Enriching the training objective with
regime-ordering ranking losses is the most direct path to making this property intrinsic.

The semantic compiler boundary --- natural language understanding in the LLM,
mathematical constraint satisfaction in the physics-informed model, auditability in
the typed DSL --- is the architectural contribution this paper makes. Each layer does
one thing well and does not ask the other layers to compensate. The next phase is
making each layer individually stronger within those boundaries: the learned model
through ranking losses, the agent through domain fine-tuning on collected trajectories,
the vocabulary layer through runtime differentiability probing. The boundaries remain;
the capabilities deepen.

% =====================================================================
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
